---
title: "Survival Models (MATH3085/6143)"
subtitle: "Chapter 1-9: Survival Models"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    progress: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: "André Victor Ribeiro Amaral"
    url: "https://www.avramaral.com/"  
    email: "a.v.ribeiro-amaral@soton.ac.uk"
    affiliations: "University of Southampton"
date: last-modified
date-format: "DD/MM/YYYY"
---

```{r setup}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
knitr::opts_chunk$set(include = TRUE)
options(digits = 4)
library("tidyverse")
```

## Chapter 1: Introduction

:::{.justify}

- [**Survival analysis**]{.secondary} refers to a set of special statistical methods required to analyse [**time-to-event**]{.alert} $T$.

<div style="margin-top: 20pt;"></div>

:::{.fragment}

- Why is Survival Analysis "special"?
  - Models for non-negative random variable $T$.
  - Data are typically [**censored**]{.alert}. 
  
:::
  
:::

## Chapter 1: Introduction ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

1. Which type of censoring occurs when an observation ends before the event of interest has been observed?
  a. Left censoring
  b. Right censoring
  c. Interval censoring
  d. This is an example of truncation
  
<div style="margin-top: 20pt;"></div>
  
:::{.fragment}

2. Which of the following is an example of non-informative censoring?
  a. A patient leaves the study because their health is rapidly improving
  b. An experimental machine is removed from the trial because it developed a new, unrecorded fault
  c. The study concludes on the final pre-scheduled date, and the subject is still alive
  d. A subject withdraws from a drug trial after experiencing severe side effects
  
:::

:::


## Chapter 2: Statistical Models

:::{.justify .small-text}

[**Parametric statistical model:**]{.secondary} for the [leukaemia survival times]{.alert}, a common model would be
$$
T_1,T_2,\cdots , T_n \stackrel{\text{i.i.d.}}{\sim} \mathrm{lognormal}(\mu,\sigma^2)
$$
where $\mu$ and $\sigma^2$ are [**unspecified**]{.alert}.

<div style="margin-top: 40pt;"></div>

:::{.fragment}

[**Non-parametric statistical model:**]{.secondary} sometimes, it is not appropriate, or we want to avoid, making a precise specification for the distribution which generated $T_1,T_2,\cdots , T_n$. 

Then, we might propose the model 
$$
T_1,T_2,\cdots , T_n \text{ are i.i.d. random variables.}
$$

<div style="margin-top: 40pt;"></div>

:::

:::{.fragment}

[**Regression model:**]{.secondary} we model survival data to learn about the relationship between survival time $T$ and other potentially explanatory variables $x_1, x_2, \cdots$.

:::

:::

## Chapter 3: The Survival distribution

:::{.justify}

- If $T$ is a continuous random variable, then its distribution is defined by its [**probability density function**]{.alert} $f_T(t)$, or $f(t)$ for short.

<div style="margin-top: -15pt;"></div>

:::{.fragment}

- The [**distribution function**]{.alert} $F_T(t)$, for a random variable $T$, is defined as $F_T(t)=\mathbb{P}(T\le t)$.

:::

<div style="margin-top: -15pt;"></div>

:::{.fragment}

- The [**survival function**]{.alert} $S_T(t)$, for a random variable $T$, is defined as $S_T(t)=\mathbb{P}(T > t)$.

:::

<div style="margin-top: -15pt;"></div>

:::{.fragment}

- The [**hazard function**]{.alert} $h_T(t)$ is defined, for $t \geq 0$, as 
$$
h_T(t) = \lim_{\delta t\to 0}\frac{\mathbb{P}(T\le t+\delta t|T> t)}{\delta t}.
$$
:::

:::{.fragment}

<div style="margin-top: -15pt;"></div>

- The [**cumulative hazard function**]{.alert} $H_T(t)$ is defined as $H_T(t)=\int_0^t h_T(u) du$.

:::

:::

## Chapter 3: The Survival distribution

:::{.justify}

- Only one of the functions $f_T$, $F_T$, $S_T$, $h_T$, or $H_T$ needs to be specified to completely determine the distribution of $T$.

<div style="margin-top: 10pt;"></div>

:::{.fragment}

[**Relationships**]{.secondary}

:::{.small-text}
| | $$f_T$$ | $$S_T$$ | $$h_T$$ |
| :---: | :---: | :---: | :---: |
| $$f_T(t) = $$ | | $$-\frac{\mathrm{d}}{\mathrm{d}t} S_T(t)$$ | $$h_T(t)\exp\left[-\int_0^t h_T(s)\mathrm{d}s\right]$$ |
| $$S_T(t) = $$ | $$\int_t^\infty f_T(s)\mathrm{d}s$$ | | $$\exp\left[-\int_0^t h_T(s)\mathrm{d}s\right]$$ |
| $$h_T(t) = $$ | $$\frac{f_T(t)}{\int_t^\infty f_T(s)\mathrm{d}s}$$ | $$-\frac{\mathrm{d}}{\mathrm{d}t} \log S_T(t)$$ | |
:::

:::

:::


## Chapter 3: The Survival distribution ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

3. The expected lifetime $\mathbb{E}(T)$ is given by
  a. The integral of $f_T(t) / S_T(t)$ over $[0, t]$
  b. The slope of the survival function
  c. The integral of the survival function, $S_T(t)$, from $0$ to $\infty$.
  d. The integral of PDF, $f_T(t)$, from $0$ to $t$.
  
<div style="margin-top: 20pt;"></div>
  
:::{.fragment}

4. How is the hazard function $h_T(t)$ related to the PDF and survival function?
  a. $h_T(t) = f_T(t) \cdot S_T(t)$
  b. $h_T(t) = -\frac{d}{dt} f_T(t)$
  c. $h_T(t) = f_T(t) / S_T(t)$
  d. $h_T(t) = S_T(t) / f_T(t)$
  
:::

:::

## Chapter 4: Distributions for Survival Models 

:::{.justify .small-text}

We introduced some [common distributions when working with time-to-event data]{.alert}, namely

- $\text{Exponential}(\beta)$: constant hazard.

:::{.fragment}

- $\text{Weibull}(\alpha, \beta)$: generalises the Exponential model; $\text{Weibull}(\alpha = 1, \beta) \equiv \text{Exponential}(\beta)$.

:::

:::{.fragment}

- $\text{Log-logistic}(\alpha, \beta)$: non-monotonic hazard.

:::

:::{.fragment}

- $\text{Log-normal}(\mu, \sigma^2)$: non-monotonic hazard (different rate of decay).

:::

:::{.fragment}

- $\text{Gompertz}(\alpha, \beta)$: risk of death increases exponentially (used to model human lifetime from middle age onwards).

:::

:::{.fragment}

- $\text{Makeham}(\alpha, \beta, \lambda)$: generalises the Gompertz model.

:::

:::

## Chapter 4: Distributions for Survival Models ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

5. If $X \sim \text{Exp}(1)$, which transformation $T = g(X)$ yields a Weibull distribution, i.e., $T \sim \text{Weibull}(\alpha, \beta)$?
  a. $T = X / \beta$
  b. $T = \exp\left\{\frac{X}{\beta}\right\}$
  c. $T = \left(\frac{X}{\beta}\right)^{1/\alpha}$
  d. $T = \frac{X^{1/ \alpha}}{\beta}$
  
:::

## Chapter 5: Survival models: parameter estimation

:::{.justify}

- In a parametric model, estimating the distribution of $T$ simply involves estimating the unknown parameters $\theta$.

:::{.fragment}

- Typically, we use [**maximum likelihood estimation**]{.alert} to obtain $\hat{\boldsymbol{\theta}}$.

:::

:::{.fragment}

- Assuming iid observations with (right) censoring indicators $d_1,\cdots ,d_n$, the likelihood function is
$$
L(\theta)=\prod_{i:d_i=1} f_T(t_i;\theta)\prod_{i:d_i=0} S_T(t_i;\theta)
$$

:::

:::{.fragment}

- For large samples, we have the asymptotic approximation
$$
\hat{\theta}\;\;\stackrel{\text{approx.}}{\sim}\; \text{Normal}(\theta, I(\theta)^{-1}).
$$

:::

:::

## Chapter 5: Survival models: parameter estimation ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

6. In the likelihood function $L(\theta)$ for a single right-censored observation $t_i$ ($d_i=0$), what is the appropriate contribution?
  a. The survival function $S_T(t_i; \theta)$
  b. The hazard function $h_T(t_i; \theta)$
  c. The distribution function $F_T(t_i; \theta)$
  d. The PDF $f_T(t_i; \theta)$
  
:::{.fragment}

7. The standard error $\text{se}(\hat{\theta})$ for the maximum likelihood estimator $\hat{\theta}$ is computed in the following way
  a. $I(\hat{\theta})^{-1}$
  b. $\sqrt{I(\hat{\theta})^{-1}}$
  c. $I(\hat{\theta})$
  d. $\sqrt{I(\hat{\theta})}$

:::
  
:::

## Chapter 6: Non-parametric Survival Estimation

:::{.justify .small-text}

Recall that the likelihood is given by

$$
L(\theta)=\prod_{i:d_i=1} f_T(t_i; \theta)\prod_{i:d_i=0} S_T(t_i; \theta).
$$

Without assuming a particular parametric family for $f_T$, the likelihood $L(\theta)$ can be made infinitely large.

To resolve this, [the **non-parametric maximum likelihood estimate** of the survival distribution is taken to be a discrete **distribution supported on the observed failure times $\{t_i: d_i=1\}$**]{.alert}.

:::{.fragment}

The [**discrete hazard MLE**]{.alert} is
$$
\hat{h}_i=\frac{d'_i}{r_i},\qquad i=1,\cdots, m, \qquad \text{ where } \qquad r_i=\sum_{j=i}^m (d'_j+c_j)
$$
is the [**number at risk**]{.alert} at $t'_i$ $(\text{failures} + \text{censored observations})$.

:::

:::

## Chapter 6: Non-parametric Survival Estimation

:::{.justify .small-text}

The corresponding [**estimator of the survival function**]{.alert}
$$
\hat{S}(t)=\prod_{j=1}^i (1-\hat{h}_j)=\prod_{j=1}^i \left(1-\frac{d'_j}{r_j}\right)\qquad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots ,m
$$
is called the [**Kaplan-Meier (or product-limit) estimator**]{.alert}.

<div style="margin-top: 20pt;"></div>

:::{.fragment}

Consider the following survival times $\mathbf{t} = \left(3, 4, 6^*, 8, 8, 10\right)$ (i.e., $m=4$ distinct observations).

<div style="margin-top: 20pt;"></div>

::: {.columns}

::: {.column width="50%"}

| $i$       | $t_i$     | $t_{i+1}$ | $r_i$    | $d_i'$    | $c_i$   | $\hat{S}(t)$ |
| :-------: | :-------: | :-------: | :-------:| :-------: |:-------:|:------------:|
| $0$ | $0$   | $3$        | $6$   | $0$    | $0$   | $1$           |
| $1$ | $3$   | $4$        | $6$   | $1$    | $0$   | $\frac{5}{6}$ |
| $2$ | $4$   | $8$        | $5$   | $1$    | $1$   | $\frac{2}{3}$ |
| $3$ | $8$   | $10$       | $3$   | $2$    | $0$   | $\frac{2}{9}$ |
| $4$ | $10$  | $\infty$   | $1$   | $1$    | $0$   | $0$           |

:::

::: {.column width="50%" .small-text}

```{r, echo = TRUE, fig.width = 6, fig.height = 3.25, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5.1, 4.1, 0.5, 2.1))
plot(0, 0, type="l", xlim=c(0,12), ylim=c(0,1), xlab = "Time", ylab = "S(time)")
segments(x0 = c(0, 3, 4, 8, 10), x1=c(3, 4, 8, 10, 12), y0=c(1, 5/6, 2/3, 2/9,0), y1 = c(1, 5/6, 2/3, 2/9, 0), lwd = 2, col = "red")
segments(x0 = c(3, 4, 8, 10),    x1=c(3, 4, 8, 10),     y0=c(1, 5/6, 2/3, 2/9  ), y1 = c(5/6, 2/3, 2/9, 0   ), lwd = 2, col = "red")
```

:::

:::

:::

:::

## Chapter 6: Non-parametric Survival Estimation ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

8. When times are tied (e.g., a failure and a censoring occur at $t^*$), the KM convention treats the censored time as
  a. Being removed entirely from the analysis
  b. Infinitesimally smaller than the failure time
  c. A full failure event ($d_i=1$)
  d. Infinitesimally larger than the failure time

:::


## Chapter 7: Survival Regression Models

:::{.justify .small-text}

A [**proportional hazards**]{.alert} model (or Cox regression model) assumes that the hazard function for the survival variable corresponding to the $i$-th unit is
$$
h_{T_i}(t)=\exp\left\{\mathbf{x}_i^{\top}\boldsymbol{\beta}\right\}\cdot h_0(t)
$$
where $h_0$ is called the [**baseline hazard function**]{.alert} and is not assumed to have a particular mathematical form.

<div style="margin-top: 20pt;"></div>

:::{.fragment}

The [**partial likelihood**]{.alert} is given by
$$
\begin{eqnarray*}
L(\boldsymbol{\beta}) = \prod_{i:d_i=1}\frac
{\exp\left(\mathbf{x}_i^{\top}\boldsymbol{\beta}\right)}
{\sum_{j\in R_i} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)}.
\end{eqnarray*}
$$

<div style="margin-top: 50pt;"></div>

:::

:::{.fragment}

An [**accelerated failure time (AFT) model**]{.alert} is generally expressed through the survival function
$$
S_{T_i}(t)=S_0\left(t \cdot \exp (-\eta_i)\right)
$$
for some [**parametric**]{.secondary} baseline survival function $S_0$. 

:::

:::

## Chapter 7: Survival Regression Models ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

9. In the Partial Likelihood denominator, the sum $\sum_{j \in R_i} \exp(\mathbf{x}_j^{\top}\boldsymbol{\beta})$ represents the
  a. Total likelihood of the uncensored observations
  b. Number of failures observed at time $t_i$
  c. Baseline hazard function $h_0(t)$
  d. Total expected risk (hazard) across all individuals in the risk set $R_i$.
  
:::{.fragment}
  
10. If the estimated coefficient $\hat{\beta}$ in a Cox model is negative $(\text{i.e., }\exp(\hat{\beta}) < 1)$, what does this imply about the covariate's effect?
  a. The covariate decreases the hazard rate (better survival)
  b. The hazard function is non-monotonic
  c. The baseline hazard $h_0(t)$ is zero
  d. The covariate increases the hazard rate

:::

:::

## Chapter 8: Multistate Survival Models

:::{.justify .small-text}

Alternatively, one can model the variable $Y_t$ representing the status (`alive` or `dead`) of a unit at time $t$.

```{r, echo = FALSE, fig.width = 8, fig.height = 3, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5.1, 4.1, 0.5, 2.1))

diff<-1.78
plot(0,0,type="n",xlab="Time",ylab="State",xlim=c(0,3),ylim=c(1,2),axes=FALSE)
segments(x0=0,x1=diff,y0=1,y1=1,lwd=2)
segments(x0=diff,x1=diff, y0=1,y1=2,lwd=2)
arrows(x0=diff,x1=3,y0=2,y1=2,len=0.1,lwd=2)
axis(side=2,at=c(1,2),labels=c("1","2"),las=2)
axis(side=1,at=(0:6)/2)
```


<div style="margin-top: -40pt;"></div>

:::{.fragment}

The [**transition probabilities**]{.alert} are defined as $\mathbb{P}(Y_{x+t}=j\,|\,Y_x=i)\;\equiv\;p_{ij}(x,t), ~~ j\in S.$

:::

:::{.fragment}

The [**transition intensity function $\mu_{ij}(x)$**]{.alert} is defined as $\mu_{ij}(x)=\lim_{\delta t\to 0} \frac{p_{ij}(x,\delta t)}{\delta t}, ~~ i\ne j.$

:::

:::{.fragment}

<div style="margin-top: 40pt;"></div>

[How are the transition probabilities $p_{ij}(x,t)$ obtained from the transition intensities $\mu_{ij}(x)$?]{.secondary}

The [**Kolmogorov forward equations**]{.alert}: $\frac{d}{dt} p_{ij}(x,t) = \sum_{k\in S} p_{ik}(x,t) \cdot \mu_{kj}(x+t), ~~ i,j\in S$.

:::

:::

## Chapter 8: Multistate Survival Models ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

11. In a time-homogeneous Markov process, the transition intensities $\mu_{ij}(x)$, $i \neq j$, are
  a. Always equal to $1$
  b. Dependent on the entire prior history of the process
  c. Dependent on the current state $i$, but not dependent on the current time $x$
  d. Always zero for all states $i$ and $j$
  
:::{.fragment}
  
12. For a time-homogeneous Markov process, the holding time distribution $T_{xi}$ is always what type of distribution?
  a. Weibull
  b. Normal
  c. Log-logistic
  d. Exponential
  
:::

:::

## Chapter 9: Inference for Multistate Models

:::{.justify .small-text}

For $n = 1$, we might observe something like

<div style="margin-top: 10pt;"></div>

| [Transition]{.alert} | Start | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | End |
| :--- | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: |
| **Time** | 0.3 | 0.32 | 0.41 | 0.56 | 0.82 | 0.84 | 1.31 | 1.41 | 1.55 | 1.79 | 2.04 | 2.24 | 2.80 |
| **State** | 2 | 1 | 3 | 4 | 1 | 2 | 1 | 3 | 2 | 3 | 1 | 2 | 2 |


:::{.fragment}

<div style="margin-top: 40pt;"></div>

The likelihood for the full set of transition intensity parameters $\boldsymbol{\mu}=\{\mu_{k\ell},k\ne\ell\}$ for a single individual transition history is

$$
\begin{eqnarray*}
L(\boldsymbol{\mu})&=&S_{y_{J}}(t_{J+1}-t_{J}) \prod_{j=1}^J f_{y_{j-1}}(t_{j}-t_{j-1})p(y_j|y_{j-1})\\ 
&=&  \prod_{k,\ell:k\ne \ell}\mu_{k\ell}^{n_{k\ell}} \cdot \exp\left(-t^+_k \mu_{k\ell}\right),
\end{eqnarray*}
$$

- $t^+_k$ is the total observed holding time in state $k$
- $n_{k\ell}$ is the total observed number of transitions from state $k$ to state $\ell$
:::

:::


## Chapter 9: Inference for Multistate Models ([vevox.app](https://vevox.app/#/m/129515844), [**129-515-844**]{.alert})

:::{.justify .small-text}

13. If the likelihood $L(\boldsymbol{\mu})$ is calculated for multiple individuals $(n > 1)$, how is the joint likelihood obtained?
  a. By averaging the individual log-likelihoods
  b. By multiplying the individual likelihoods, assuming independence
  c. By using numerical methods only
  d. By summing the individual likelihoods

:::