---
title: "Survival Models (MATH3085/6143)"
subtitle: "Chapter 10: Modelling Human Lifetime"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    progress: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: "André Victor Ribeiro Amaral"
    url: "https://www.avramaral.com/"  
    email: "a.v.ribeiro-amaral@soton.ac.uk"
    affiliations: "University of Southampton"
date: last-modified
date-format: "DD/MM/YYYY"
---

```{r setup}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
knitr::opts_chunk$set(include = TRUE)
options(digits = 4)
library("tidyverse")
```

## Recap

:::{.justify}

In the last chapter, we

- Derived the likelihood for an individual transition history and showed that the [**MLE for the time-homogeneous intensity**]{.alert} is simply the total number of transitions from $k$ to $\ell$ divided by the total observed holding time in $k$, i.e.,
$$
\hat{\mu}_{k\ell} = \frac{n_{k\ell}}{t^+_k}, \qquad k \neq \ell.
$$
- Derived the [**standard errors**]{.alert} for $\hat{\mu}_{k\ell}$ and constructed [**confidence intervals**]{.alert} around our estimates.

- Generalised the [**likelihood for multiple histories of transitions**]{.alert} corresponding to individuals $i = 1, \cdots, n$.

:::

# <span style="font-size: 42px; color: #131516; display: block; margin-bottom: -50px">[**Chapter 10:**]{.secondary} Modelling Human Lifetime</span>

## Introduction

:::{.justify .small-text}

For the rest of this module our focus will be on modeling [**human life length**]{.alert}, that is the time (from birth or some other specified time origin) to death in a specified human population.

:::{.fragment}

<div style="margin-top: 30pt;"></div>

[Models for human human lifetimes have a huge importance]{.alert}, for example

<div style="margin-top: 20pt;"></div>

- Billions of pounds are invested in pension funds. [Calculation of the liabilities]{.secondary} requires us to be able to predict lifetimes of current and future pensioners.

:::

:::{.fragment}

- Planning public services requires requires us to predict age-structured populations, sometimes within a small geographical area. This requires us to be able to [forecast mortality]{.secondary} (along with fertility and migration).

:::

:::


## Special features of human lifetime

:::{.justify .small-text}

There are a number of practical reasons why human lifetime modelling requires [**special treatment**]{.alert}, e.g.,

- The [**long time scales involved**]{.secondary} (much longer than the horizon of a standard statistical study).

:::{.fragment}


- The requirement to [**use secondary data**]{.secondary} (collected for other purposes, such as census and death registration data)

:::

:::{.fragment}

- Data sets are often large (`good`) but the data can be coarse (`bad`). For example, ages may only be provided in whole years.

:::

:::{.fragment}

- Standard distributions tend not to provide a good fit for human lifetimes.

:::

:::{.fragment}

- The distribution of human lifetime is changing (lifetimes are getting longer).

:::

:::

## Models

:::{.justify .small-text}

As previously, we use $T$ to denote the [**time**]{.alert} from a specified origin (usually birth) [**until death**]{.alert}.

Then we know that a survival model can be specified by the survival function $S_T(t)$ or the hazard function $h_T(t)$ for $T$.

:::{.fragment}

Alternatively, we can think of human survival as a two-state process for $Y_x\in\{1,2\}$ (`alive` and `dead`) with an absorbing state.

```{r, echo = FALSE, fig.width = 10, fig.height = 2, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(0, 0, 0, 0))

plot(0,0,xlim=c(0,100),ylim=c(0,40),type="n",axes=FALSE,xlab="",ylab="")

polygon(x=c(20,20,40,40),y=c(10,30,30,10))
polygon(x=c(20,20,40,40)+40,y=c(10,30,30,10))
arrows(x0=40,x1=60,y0=20,y1=20,len=0.1)
text(x=30,y=20,labels=c("Alive"),pos=3)
text(x=30,y=20,labels=c("(Y=1)"),pos=1)
text(x=30+40,y=20,labels=c("Dead"),pos=3)
text(x=30+40,y=20,labels=c("(Y=2)"),pos=1)
```

<div style="margin-top: -30pt;"></div>

Then the process is specified by the transition intensity function $\mu_{12}(x)$ at time $x$ or, alternatively, the transition probabilities $p_{11}(x,t)=1-p_{12}(x,t)$.

:::

:::

## Models are equivalent [**(Gap on page 114)**]{.alert}

:::{.justify .small-text .fragment}

Recall that 

$$
\begin{eqnarray*}
p_{11}(x,t)\;=\; P(Y_{x+t}=1\,|\,Y_x=1)&=&P(T> x+t\,|\,T>x)\\[2pt]
&=&S_T(t|x)\\[2pt]
&=&\frac{S_T(x+t)}{S_T(x)},
\end{eqnarray*}
$$
where $S_T(t|x)$ is the residual survival function (see [**Section 3.3.2**]{.secondary}). So, $p_{11}(0,t)=S_T(t)$

<div style="margin-top: 20pt;"></div>

Similarly,
$$
\mu_{12}(x)=\lim_{\delta t\to 0} \frac{p_{12}(x,\delta t)}{\delta t}
=\lim_{\delta t\to 0} \frac{P(T\le  x+\delta t\,|\,T>x)}{\delta t } =h_T(x).
$$

So the models are equivalently specified.

:::

## Alternative notation [**(Gap on page 115)**]{.alert}

:::{.justify .small-text .fragment}

Because of the importance of [**human lifetime modelling in demography and actuarial science**]{.secondary}, a specific notation has been developed.

<div style="margin-top: 30pt;"></div>

:::{.fragment}

$$
p_{11}(x,t)=S_T(t|x)\;\equiv\;\,_tp_x
$$


:::

<div style="margin-top: 30pt;"></div>

:::{.fragment}

and
$$
p_{12}(x,t)=1-S_T(t|x)=1-\,_tp_x\;\equiv\;\,_tq_x.
$$

<div style="margin-top: 30pt;"></div>

:::

:::{.fragment}

So, 

- ${}_tq_x$ is the probability of death in $(x,x+t]$ given survival to age $x$.
- ${}_tp_x$ is probability of probability of survival to $x+t$ given survival to age $x$.

:::

:::


## $p_x$ and $q_x$ notation [**(Gap on page 116)**]{.alert}

:::{.justify .small-text .fragment}

When $t=1$, the first subscript is usually omitted, so

- $q_x$ is the probability of death in $(x,x+1]$ given survival to age $x$
- $p_x= 1 - q_x$ is probability of survival to $x+1$ given survival to age $x$.

:::{.fragment}

This is most commonly used when $x$ is an exact integer number of years.

:::

:::{.fragment}

Note that if $t$ is an integer then
$$
\begin{eqnarray*}
{}_tp_x &=\frac{S_T(x+t)}{S_T(x)} &= \frac{S_T(x+1)}{S_T(x)} \times \frac{S_T(x+2)}{S_T(x+1)} \times \cdots \times \frac{S_T(x+t)}{S_T(x+t-1)}\\[10pt]
&&=p_{x}\times p_{x+1}\times \cdots \times p_{x+t-1}
\end{eqnarray*}
$$

:::

:::{.fragment}

and so
$$
{}_tq_x=1-(1-q_{x})\times (1-q_{x+1}) \times \cdots \times (1-q_{x+t-1}).
$$

:::

:::

## Force of mortality [**(Gap on page 117)**]{.alert}

:::{.justify .small-text .fragment}

We also have that 
$$
\mu_{12}(x)=h_T(x)\;\equiv\;\mu_x,
$$
which is called [(in human lifetime applications)]{.secondary} the [**force of mortality**]{.alert} at age $x$.

<div style="margin-top: 20pt;"></div>

:::{.fragment}

The force of mortality $\mu_x$ is the limiting death rate at age $x$, and as such is only constrained by $\mu_x\ge 0$.

:::

:::{.fragment}

In practice, if we are measuring time in years, $\mu_x<1$ for all except [**very**]{.alert} extreme ages $x$.

:::

<div style="margin-top: 20pt;"></div>

:::{.fragment}

The functions ${}_tp_x$, ${}_tq_x$ and $\mu_x$ are related by
$$
{}_tp_x\;=\;1-{}_tq_x\;=\;\exp\left(-\int_x^{x+t} \mu_x \mathrm{d}x\right).
$$

:::
:::