---
title: "Survival Models (MATH3085/6143)"
subtitle: "Chapter 6: Non-parametric Survival Estimation"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    progress: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: "André Victor Ribeiro Amaral"
    url: "https://www.avramaral.com/"  
    email: "a.v.ribeiro-amaral@soton.ac.uk"
    affiliations: "University of Southampton"
date: last-modified
date-format: "DD/MM/YYYY"
---

```{r setup}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
knitr::opts_chunk$set(include = TRUE)
options(digits = 4)
library("tidyverse")
library("survival")	
library("MASS")
```

## Recap

:::{.justify}

In the last chapter, we

- Defined the likelihood function $L(\theta)$, showing how to [**incorporate all censoring types**]{.alert} (although we will focus on right-censored data only).

- Identified [**Maximum Likelihood Estimation (MLE)**]{.alert} as the primary method for finding the parameter estimates $\hat{\theta}$ in parametric survival models by maximizing the log-likelihood function, $\ell(\theta)$.

- Derived the MLE for the Exponential model.

- Estimated parameters for more complex models (e.g., Weibull) using [**numerical methods**]{.alert}. We also compared the results with `flexsurvreg()` (from `flexsurv`).

:::

# <span style="font-size: 42px; color: #131516; display: block; margin-bottom: -50px">[**Chapter 6:**]{.secondary} Non-parametric Survival Estimationn</span>


## Introduction

:::{.justify .small-text}

Frequently, we want to estimate the distribution of $T$ for the  (i.i.d.) homogeneous model based on  

<div style="margin-top: -10pt;"></div>

- observations $t_1, \cdots ,t_n$ 
- with (right-) censoring indicators $d_1, \cdots ,d_n$, 

<div style="margin-top: -25pt;"></div>

[**without**]{.alert} assuming a particular parametric family for $f_T$.

<div style="margin-top: 60pt;"></div>

:::{.fragment}
The likelihood is given by 
$$
L(\theta)=\prod_{i:d_i=1} f_T(t_i; \theta)\prod_{i:d_i=0} S_T(t_i; \theta)
$$
and this can be made infinitely large by concentrating $f_T$ in infinitesimally narrow regions around each observed $t_i$ (i.e. those for which  $d_i=1$). That is, if we only require $f_T(t) \ge 0$ for all $t \in \mathbb{R}_+$ and $\int_0^{\infty}f_T(t)dt = 1$, without any additional smoothness or parametric constraints, then [the likelihood is unbounded]{.alert}.


To resolve this, [the **non-parametric maximum likelihood estimate** of the survival distribution is taken to be a discrete **distribution supported on the observed failure times $\{t_i: d_i=1\}$**]{.alert}.

:::
:::


## Discrete survival distributions [**(Gap on page 45)**]{.alert}

:::{.justify .small-text .fragment}

A discrete survival variable $T$ takes values in a sample space $0<t'_1<t'_2<\cdots <t'_m<\infty$ and is defined by any of

<div style="margin-top: 35pt;"></div>

:::{.fragment}
- Probability function 

$$
f_i =\mathbb{P}(T=t'_i), \qquad \text{for}~~ i=1, \cdots, m.
$$

<div style="margin-top: 35pt;"></div>

:::

:::{.fragment}
- Survival function

$$
\begin{eqnarray*}
S(t)=\mathbb{P}(T>t)&=&1-\sum_{j:t'_j\le t} f_j\\
&\equiv& s_i, \qquad \text{for}~~ t\in[t'_i,t'_{i+1}) ~~\text{and}~~ i=0, \cdots ,m,
\end{eqnarray*}
$$
where $t'_0 \equiv 0$ and $t'_{m+1} \equiv \infty$.
:::
:::

## Discrete survival distributions [**(Gap on page 45)**]{.alert}

:::{.justify .small-text}

- Hazard function  

<div style="margin-top: -15pt;"></div>

$$
\begin{eqnarray*}
h_i & = & \mathbb{P}(T=t'_i|T\ge t'_i) = \frac{\mathbb{P}(T = t'_i \quad \mbox{and} \quad T\ge t'_i)}{\mathbb{P}(T\ge t'_i)} =  \frac{\mathbb{P}(T = t'_i)}{\mathbb{P}(T\ge t'_i)} = \frac{f_i}{\mathbb{P}(T > t'_{i-1})}\\
& = & \frac{f_i}{s_{i-1}}
\end{eqnarray*}
$$

<div style="margin-top: -15pt;"></div>

We have $s_0=1$ and

$$
\begin{eqnarray*}
s_i & = & \mathbb{P}(T > t_i') =  1 - \mathbb{P}(T \le t_i') =  1 - \left(\mathbb{P}(T \le t_{i-1}') + f_i \right)\\
& =& \mathbb{P}(T>t_{i-1}') - f_i = s_{i-1} - f_i =  s_{i-1}-h_i s_{i-1} =  s_{i-1}(1-h_i )
\end{eqnarray*}
$$
<div style="margin-top: -15pt;"></div>

Therefore,
$$
s_i=\prod_{j=1}^i (1-h_j)\qquad{\rm and}\qquad f_i=h_i\prod_{j=1}^{i-1} (1-h_j)
$$

It is [**best to describe the survival distributions through $\{h_i\}$**]{.alert}, which are simply constrained to be in $[0,1]$, rather than $\{f_i\}$ or $\{s_i\}$ (which have more complex constraints).

:::

## Likelihood [**(Gap on page 48)**]{.alert}

:::{.justify .small-text}

We will now [connect these functions in the discrete case to the likelihood]{.alert} discussed in the previous chapter. To do so,

<div style="margin-top: 15pt;"></div>

:::{.fragment}

- Let $0<t'_1<t'_2<\cdots <t'_m<\infty$ denote the [**distinct**]{.alert} observed failure times observed in our sample $\{t_1,\cdots , t_n\}$.

<div style="margin-top: 25pt;"></div>

- $t'_0=0$ and $t'_{m+1}=\infty$.

<div style="margin-top: 25pt;"></div>

- $d'_i$ is the number of failures observed at $t'_i,\; ( i=1,\cdots ,m)$, so

$$
\sum_{i=1}^m  d'_i = \sum_{i=1}^n  d_i=d_+.
$$

<div style="margin-top: 15pt;"></div>

- $c_i\; ( i=0,\cdots ,m)$ is the number of censored observations with censoring times between $t'_i$ and $t'_{i+1}$.

:::
:::

## Likelihood [**(Gap on page 48)**]{.alert}

:::{.justify .small-text}

The [**likelihood**]{.alert} is given by 

$$
\begin{eqnarray*}
L(\theta)=\prod_{i=1}^m f_i^{d'_i} \prod_{i=0}^m s_i^{c_i}
&=&\prod_{i=1}^m \left[h_i\prod_{j=1}^{i-1} (1-h_j)\right]^{d'_i} \prod_{i=1}^m \left[\prod_{j=1}^i (1-h_j)\right]^{c_i}\\
&=&\prod_{i=1}^m  \left[\left(\frac{h_i}{1-h_i}\right)^{d'_i}\prod_{j=1}^i (1-h_j)^{d'_i+c_i}\right]\\
&=&\left[\prod_{i=1}^m \left(\frac{h_i}{1-h_i}\right)^{d'_i}\right]\prod_{i=1}^m\prod_{j=1}^i (1-h_j)^{d'_i+c_i}\\
&=&\left[\prod_{i=1}^m \left(\frac{h_i}{1-h_i}\right)^{d'_i}\right]\prod_{i=1}^m\prod_{j=i}^m (1-h_i)^{d'_j+c_j}\\
&=&\left[\prod_{i=1}^m \left(\frac{h_i}{1-h_i}\right)^{d'_i}\right]\prod_{i=1}^m(1-h_i)^{\sum_{j=i}^m (d'_j+c_j)}\\
\end{eqnarray*}.
$$
:::

## Likelihood [**(Gap on page 48)**]{.alert}

:::{.justify .small-text}

The [**log-likelihood**]{.alert} is given by 

$$
\ell(\theta)=\sum_{i=1}^m \left[d'_i\log h_i -d'_i\log(1-h_i) +\log (1-h_i)\sum_{j=i}^m (d'_j+c_j)\right],
$$

so

$$
\frac{\partial\ell}{\partial h_i}=\frac{d'_i}{h_i}+ \frac{d'_i}{1-h_i} - \frac{\sum_{j=i}^m (d'_j+c_j)}{1-h_i}.
$$
and

$$
\begin{eqnarray*}
&&\frac{d'_i}{\hat{h}_i}+ \frac{d'_i-\sum_{j=i}^m (d'_j+c_j)}{1-\hat{h}_i} ~~\overset{\text{SET}}{=}~~ 0\\
&\text{implying}\qquad&\hat{h}_i=\frac{d'_i}{\sum_{j=i}^m (d'_j+c_j)},\qquad i=1,\cdots,m.
\end{eqnarray*}
$$

:::

## The Kaplan-Meier estimator [**(Gap on page 49)**]{.alert}

:::{.justify .small-text .fragment}

The [**discrete hazard MLE**]{.alert} is
$$
\hat{h}_i=\frac{d'_i}{r_i},\qquad i=1,\cdots, m, \qquad \text{ where } \qquad r_i=\sum_{j=i}^m (d'_j+c_j)
$$
is the [**number at risk**]{.alert} (of failure) at $t'_i$.
 
The hazard at each $t'_i$ is therefore estimated by the observed number of failures at $t'_i$ as a proportion of the number at risk at  $t'_i$.

<div style="margin-top: 40pt;"></div>

:::{.fragment}
The corresponding [**estimator of the survival function**]{.alert}
$$
\hat{S}(t)=\prod_{j=1}^i (1-\hat{h}_j)=\prod_{j=1}^i \left(1-\frac{d'_j}{r_j}\right)\qquad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots ,m
$$
is called the [**Kaplan-Meier (or product-limit) estimator**]{.alert}.
:::
:::

## Kaplan-Meier estimator 

:::{.justify .small-text}


[**Gehan data**]{.alert}: remission times (in weeks) from a clinical trial of 42 leukaemia patients. Patients matched in pairs and randomised to 6-mercaptopurine (drug) or control (from `MASS` package).

<div style="margin-top: -15pt;"></div>

Using the `survfit()` and `Surv()` functions from the `survival` package, we can compute the KM curves as follows

<div style="margin-top: -5pt;"></div>

```{r, echo = TRUE, fig.width = 6, fig.height = 2.25, fig.align = "center"}
gehanControl <- survfit(Surv(time = gehan$time[gehan$treat == "6-MP"   ], event = gehan$cens[gehan$treat == "6-MP"]   ) ~ 1)
gehan6mp     <- survfit(Surv(time = gehan$time[gehan$treat == "control"], event = gehan$cens[gehan$treat == "control"]) ~ 1)

par(family = "Latin Modern Roman 10", mar = c(1.85, 4.1, 0.5, 2.1))
plot( gehan6mp,     mark.time = TRUE, conf.int = FALSE, col = "blue", lwd = 2, xlab = "Time (weeks)", ylab = "Survival", main = NA, xlim = c(0, 35))
lines(gehanControl, mark.time = TRUE, conf.int = FALSE, col = "red",  lwd = 2)
legend("topright", legend = c("6-MP", "Control"), col = c("blue", "red"), lwd = 2, bty = "n")
# Alternatively, one can use `treat` as a covariate. This would produce the same plot.
# gehan.km <- survfit(Surv(time, cens) ~ treat, data = gehan)
# plot(gehan.km, xlab = "Time (weeks)", ylab = "S(time)", lty = c(1, 2), xlim = c(0, 35))

```


:::


## Notes on Kaplan-Meier estimator [**(Gap on page 50)**]{.alert}

:::{.justify .fragment}

- Although the Kaplan-Meier estimator represents a discrete survival distribution, we use it to estimate $S_T(t)$ for [continuous $T$]{.alert}.

<div style="margin-top: 5pt;"></div>

:::{.fragment}

- For censored times tied with uncensored times, we treat the censored times as infinitessimally
larger than the uncensored ones with which they are tied (so $c_i$ includes observations censored at $t'_i$).

<div style="margin-top: 5pt;"></div>

:::

:::{.fragment}

- If the largest time is censored (so $c_m>0$), then the Kaplan-Meier estimator is positive at the largest uncensored time and therefore the estimated survivor function is [nowhere zero]{.alert}.

<div style="margin-top: 5pt;"></div>

:::

:::{.fragment}

- Sometimes, it is informative to plot $\log{S}(t)$, recalling that the hazard function $h(t)$ is equal to $-\frac{\mathrm{d}}{\mathrm{d}t}\log S(t)$.
Hence $\log{S}(t)$ is linear for [exponentially distributed data (constant hazard)]{.alert}.

:::
:::

## Kaplan-Meier estimate by hand [**(Gap on page 50)**]{.alert}

:::{.justify}

Consider the following survival times $\mathbf{t} = \left(3, 4, 6^*, 8, 8, 10\right)$. There are $m=4$ distinct observed survival times. 

:::{.fragment}

<div style="margin-top: 50pt;"></div>

::: {.columns}

::: {.column width="50%"}

| $i$       | $t_i$     | $t_{i+1}$ | $r_i$    | $d_i'$    | $c_i$   | $\hat{S}(t)$ |
| :-------: | :-------: | :-------: | :-------:| :-------: |:-------:|:------------:|
| $0$ | $0$   | $3$        | $6$   | $0$    | $0$   | $1$           |
| $1$ | $3$   | $4$        | $6$   | $1$    | $0$   | $\frac{5}{6}$ |
| $2$ | $4$   | $8$        | $5$   | $1$    | $1$   | $\frac{2}{3}$ |
| $3$ | $8$   | $10$       | $3$   | $2$    | $0$   | $\frac{2}{9}$ |
| $4$ | $10$  | $\infty$   | $1$   | $1$    | $0$   | $0$           |

:::

::: {.column width="50%" .small-text}

```{r, echo = TRUE, fig.width = 6, fig.height = 3.25, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5.1, 4.1, 0.5, 2.1))

plot(0, 0, type="l", xlim=c(0,12), ylim=c(0,1), xlab = "Time", ylab = "S(time)")
segments(x0 = c(0, 3, 4, 8, 10), x1=c(3, 4, 8, 10, 12), y0=c(1, 5/6, 2/3, 2/9,0), y1 = c(1, 5/6, 2/3, 2/9, 0), lwd = 2, col = "red")
segments(x0 = c(3, 4, 8, 10),    x1=c(3, 4, 8, 10),     y0=c(1, 5/6, 2/3, 2/9  ), y1 = c(5/6, 2/3, 2/9, 0   ), lwd = 2, col = "red")
```

:::
:::
:::
:::

## Kaplan-Meier with `R` 

:::{.justify .small-text}

As before, we can compute it in `R` using the `survival` package.

<div style="margin-top: 10pt;"></div>

```{r, echo = TRUE, fig.width = 9, fig.height = 3.55, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5.1, 4.1, 2.5, 2.1))

time <- c(3, 4, 6, 8, 8, 10)
cens <- c(1, 1, 0, 1, 1, 1 )

km <- survfit(Surv(time, cens) ~ 1)

par(mfrow=c(1,2))
plot(km, conf.int = FALSE, xlab = "Time", ylab = "S(time)", main = "Without 'censoring marks' and CI")
plot(km, mark.time = TRUE, xlab = "Time", ylab = "S(time)", main =    "With 'censoring marks' and CI")

```
:::

## Standard Errors and CIs [**(Gap on page 55)**]{.alert}

:::{.justify .fragment}

The standard error of the Kaplan-Meier estimator is given by [[**Greenwood's formula**]{.alert} (i.e., a formula used to approximate the variance of the Kaplan-Meier estimator)](https://en.wikipedia.org/wiki/Kaplan%E2%80%93Meier_estimator#Statistical_considerations){target="_blank"}
$$
\text{se}[\hat{S}(t)]^2=\hat{S}(t)^2
\sum_{j=1}^i {{d'_j}\over{r_j(r_j-d'_j)}}\qquad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots ,m
$$

Using a normal approximation for the sampling distribution of $\hat{S}(t)$, we can obtain endpoints of an approximate $(1-\alpha)\%$ confidence interval for $S(t)$ as 
$$
\hat{S}(t)\pm z_{\left(1-\frac{\alpha}{2}\right)} \cdot \text{se}[\hat{S}(t)].
$$


:::

## Standard Errors and CIs [**(Gap on page 55)**]{.alert}

:::{.justify}
Better confidence intervals are often obtained on the log scale, using
$$
\text{se}[\log\hat{S}(t)]^2=
\sum_{j=1}^i {{d'_j}\over{r_j(r_j-d'_j)}}\qquad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots ,m
$$
to give endpoints of an alternative $(1-\alpha)\%$ confidence interval as
$$
\hat{S}(t)\cdot \exp\left\{\pm z_{\left(1-\frac{\alpha}{2}\right)}\cdot \text{se}[\log\hat{S}(t)]\right\}.
$$
:::


## The Nelson-Aalen estimator [**(Gap on page 56)**]{.alert}

:::{.justify .fragment}

The [**Nelson-Aalen estimator**]{.alert} estimates the cumulative hazard function $H(t)$ by integrating (summing) the discrete hazard estimators $\hat{h}_i=d'_i/r_i$, and is given by

:::{.small-text}
$$
\hat{H}(t)=\sum_{j=1}^i \hat{h}_j=\sum_{j=1}^i \frac{d'_j}{r_j}\qquad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots ,m.
$$
:::

<div style="margin-top: 20pt;"></div>

:::{.fragment}
As $S(t)=\exp \{-H(t)\}$, [the Nelson-Aalen estimator provides an estimator of $S(t)$]{.alert} as

:::{.small-text}
$$
\hat{S}(t)= \exp\left(-\sum_{j=1}^i\frac{d'_j}{r_j}\right)= \prod_{j=1}^i\exp\left(-\frac{d'_j}{r_j}\right)\qquad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots,m
$$
:::
which is sometimes called the [**Fleming-Harrington estimator**]{.alert}.

:::
:::

## Notes on Nelson-Aalen estimator [**(Gap on page 57)**]{.alert}

:::{.justify .fragment .small-text}

Plotting $\log[{\hat H}(t)]$ against $\log(t)$ checks whether a Weibull model might fit the data, as for the Weibull,
$$
H(t)=(\beta t)^\alpha\qquad\Rightarrow\qquad \log [H(t)]=\alpha\log(\beta)+\alpha\log(t),
$$
so for Weibull data, we would expect $\log {\hat H}(t)$ to be linear in $\log(t)$, with gradient $\alpha$.

<div style="margin-top: 40pt;"></div>

:::{.fragment}
If $d'_i/r_i$ is small (typical for smaller values of $t$) then the Kaplan-Meier and Fleming-Harrington estimators of $S(t)$ will be close, as
$$
\begin{eqnarray*}
exp\left(-\frac{d'_j}{r_j}\right)&=&1-\frac{d'_j}{r_j}+\frac{1}{2}\left(\frac{d'_j}{r_j}\right)^2 -\cdots\\
&\approx& 1-\frac{d'_j}{r_j}.
\end{eqnarray*}
$$
:::
:::