---
title: "Survival Models (MATH3085/6143)"
subtitle: "Chapter 9: Inference for Multistate Models"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    progress: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: "André Victor Ribeiro Amaral"
    url: "https://www.avramaral.com/"  
    email: "a.v.ribeiro-amaral@soton.ac.uk"
    affiliations: "University of Southampton"
date: last-modified
date-format: "DD/MM/YYYY"
---

```{r setup}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
knitr::opts_chunk$set(include = TRUE)
options(digits = 4)
library("tidyverse")
```

## Recap

:::{.justify}

In the last chapter, we

- We moved from modelling the time-to-event variable $T$ to modelling the state variable $Y_t$, representing a unit's status over time, using the framework of a [**stochastic process**]{.alert}.

- Defined the [**transition intensity function $\mu_{ij}(x)$**]{.alert} as the multi-state equivalent of the hazard function. This instantaneous rate is the core parameter used to define movement from state $i$ to state $j$.

- Derived the [**Kolmogorov Forward Equations**]{.alert} to obtain the [transition probabilities $p_{ij}(x,t)$]{.secondary} from the transition intensity function $\mu_{ij}(x)$.

:::

# <span style="font-size: 42px; color: #131516; display: block; margin-bottom: -50px">[**Chapter 9:**]{.secondary} Inference for Multistate Models</span>

## Introduction

:::{.justify .small-text}


In [**Chapter 8**]{.secondary}, we have introduced multistate models and their properties, and presented several examples of models with possible applications to survival data analysis.

:::{.fragment}

A key part of the statistical analysis process is [**making inference about the parameters of the model**]{.alert}, on the basis of observed data.

:::

:::{.fragment}

For a multistate model, with states $1,\cdots ,m$ the parameters are the transition intensity functions
$$
\mu_{k\ell}(x),\quad i,j=1,\cdots , m,\quad k\ne\ell.
$$

:::

:::{.fragment}

[We will only consider time-homogenous models]{.alert}, so the parameters to be estimated are the 
transition intensities
$$
\mu_{k\ell},\quad i,j=1,\cdots , m,\quad k\ne\ell.
$$

:::
:::


## Data

:::{.justify .small-text}

For a multistate model, the data will be the transition histories of a set of individuals $i=1,\cdots ,n$. 

:::{.fragment}

We start by considering the case $n=1$.

<div style="margin-top: 10pt;"></div>

In this case, [**we might observe something like**]{.secondary}

| [Transition]{.alert} | Start | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | End |
| :--- | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: |
| **Time** | 0.3 | 0.32 | 0.41 | 0.56 | 0.82 | 0.84 | 1.31 | 1.41 | 1.55 | 1.79 | 2.04 | 2.24 | 2.80 |
| **State** | 2 | 1 | 3 | 4 | 1 | 2 | 1 | 3 | 2 | 3 | 1 | 2 | 2 |


```{r, echo = FALSE, fig.width = 10, fig.height = 3.5, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5, 5, 3, 3))

trans<-0:12
state<-c(2,1,3,4,1,2,1,3,2,3,1,2,2)
time<-c(0.3,0.32,0.41,0.56,0.82,0.84,1.31, 1.41,1.55,1.79,2.04,2.24,2.80)

plot(0,0,xlab="Time",ylab="State",axes = FALSE, type="n", xlim=c(0,3), ylim=c(0.5,4.5))
axis(side=2,at=1:4,labels=c("1","2","3","4"),las=2)
axis(side=1,at=(0:6)/2)
for(i in 1:12){
  segments(x0=time[i],x1=time[i+1],y0=state[i],lwd=2, col = "red")
}
```

:::

:::

## Data structure [**(Gap on page 105)**]{.alert}

:::{.justify .small-text}


We will denote the data observations using

<div style="margin-top: 5pt;"></div>

| Transition ($j$) | 0 | 1 | 2 | $\cdots$ | $j$ | $\cdots$ | $J$ | $J+1$ (end) |
| :----- | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: |
| **Time ($t_j$)** | $t_0$ | $t_1$ | $t_2$ | $\cdots$ | $t_j$ | $\cdots$ | $t_J$ | $t_{J+1}$ |
| **State ($y_j$)** | $y_0$ | $y_1$ | $y_2$ | $\cdots$ | $y_j$ | $\cdots$ | $y_J$ | |


<div style="margin-top: 35pt;"></div>

:::{.fragment}

The [**likelihood**]{.alert} for the transition intensity parameters $\boldsymbol{\mu}=\{\mu_{k\ell},k\ne\ell\}$ is the probability of the observed sequence of holding times and transitions, given $\boldsymbol{\mu}$, that is

$$
\begin{eqnarray*}
L(\boldsymbol{\mu})&=&f_{y_0}(t_1-t_0)p(y_1|y_0)\;\times\;
f_{y_1}(t_2-t_1)p(y_2|y_1)\; \times\;\cdots\\ &&\qquad \times\;
f_{y_{J-1}}(t_{J}-t_{J-1})p(y_J|y_{J-1})\;\times\; S_{y_{J}}(t_{J+1}-t_{J}),
\end{eqnarray*}
$$

where 

- $f_k(t)$ is the exponential probability density function of the holding time $t$ in state $k$ and 
- $p(\ell|k)$ is the conditional probability of transition from state $k$ to state $\ell$ given that a transition has happened.

:::

:::

## Likelihood [**(Gap on page 106)**]{.alert}

:::{.justify .small-text}

To compute the [**likelihood**]{.alert}, we have


:::{.fragment}
$$
\begin{eqnarray*}
L(\boldsymbol{\mu})&=&f_{y_0}(t_1-t_0)p(y_1|y_0)\;\times\;
f_{y_1}(t_2-t_1)p(y_2|y_1)\; \times\;\cdots\\ &&\qquad \times\;
f_{y_{J-1}}(t_{J}-t_{J-1})p(y_J|y_{J-1})\;\times\; S_{y_{J}}(t_{J+1}-t_{J})\\
&=&
S_{y_{J}}(t_{J+1}-t_{J}) \prod_{j=1}^J f_{y_{j-1}}(t_{j}-t_{j-1}) \cdot p(y_j|y_{j-1}),
\end{eqnarray*}
$$
where 
$$
f_k(t)=\sum_{\ell\ne k} \mu_{k\ell}\exp\left(-t \sum_{\ell\ne k} \mu_{k\ell}\right),
$$
$$
S_k(t)=\exp\left(-t \sum_{\ell\ne k} \mu_{k\ell}\right), ~~~~\text{and}
$$
$$
p(\ell|k)=\frac{\mu_{k\ell}}{\sum_{\ell\ne k} \mu_{k\ell}}.
$$


:::

:::


## Likelihood [**(Gap on page 106)**]{.alert}

:::{.justify .small-text}

Hence,
$$
\begin{eqnarray*}
L(\boldsymbol{\mu})&=&S_{y_{J}}(t_{J+1}-t_{J}) \prod_{j=1}^J f_{y_{j-1}}(t_{j}-t_{j-1})p(y_j|y_{j-1})\\
&=&\exp\left(-(t_{J+1}-t_{J})\sum_{\ell\ne y_J} \mu_{y_J\ell}\right)\\ 
&&\qquad\times \prod_{j=1}^J \exp\left(-(t_{j}-t_{j-1})
 \sum_{\ell\ne y_{j-1}} \mu_{y_{j-1}\ell}\right)  \mu_{y_{j-1}y_j}\\
 &=&\prod_{j=1}^{J+1} \exp\left(-(t_{j}-t_{j-1})
 \sum_{\ell\ne y_{j-1}} \mu_{y_{j-1}\ell}\right)  \prod_{j=1}^J \mu_{y_{j-1}y_j}\\
&=&\prod_{j=1}^{J+1} \prod_{\ell\ne y_{j-1}} \exp\left\{-(t_{j}-t_{j-1})
  \mu_{y_{j-1}\ell}\right\}  \prod_{j=1}^J \mu_{y_{j-1}y_j}.
\end{eqnarray*}
$$

So the likelihood factorises into a series of terms, each containing a single transition intensity $\mu_{k\ell}$.

:::


## Likelihood [**(Gap on page 106)**]{.alert}

:::{.justify .small-text}

Gathering together all the terms containing a particular $\mu_{k\ell}$, we obtain
$$
\begin{eqnarray*}
L(\mu_{k\ell})&=&\mu_{k\ell}^{n_{k\ell}}
\prod_{j:y_{j-1}=k}^{J+1} \exp\left\{-(t_{j}-t_{j-1})\mu_{k\ell}\right\}  \\
&=&\mu_{k\ell}^{n_{k\ell}}\exp\left(-t^+_k \mu_{k\ell}\right) 
\end{eqnarray*}
$$
where 

- $t^+_k$ is the total observed holding time in state $k$
- $n_{k\ell}$ is the total observed number of transitions from state $k$ to state $\ell$

are called [**sufficient statistics**]{.secondary}.

:::{.fragment}

Therefore the  likelihood for the full set of transition intensity parameters $\boldsymbol{\mu}=\{\mu_{k\ell},k\ne\ell\}$ for a single individual transition history is
$$
L(\boldsymbol{\mu})=\prod_{k,\ell:k\ne \ell}\mu_{k\ell}^{n_{k\ell}} \cdot \exp\left(-t^+_k \mu_{k\ell}\right).
$$

:::

:::

## MLE for $\mu$ [**(Gap on page 109)**]{.alert}

:::{.justify .small-text .fragment}

The log likelihood is
$$
\ell(\boldsymbol{\mu})=\sum_{k,\ell:k\ne \ell}{n_{k\ell}}\log \mu_{k\ell} -\sum_{k,\ell:k\ne \ell}t^+_k \mu_{k\ell} 
$$
and therefore
$$
\frac{\partial}{\partial\mu_{k\ell}} \ell(\boldsymbol{\mu}) 
=\frac{n_{k\ell}}{\mu_{k\ell}} -t^+_k 
$$
so the MLE solves 

$$
\frac{n_{k\ell}}{\hat\mu_{k\ell}} -t^+_k = 0 \qquad\Rightarrow\qquad \hat\mu_{k\ell}=\frac{n_{k\ell}}{t^+_k}.
$$

:::

## Standard errors for $\hat{\mu}$ [**(Gap on page 109)**]{.alert}

:::{.justify .small-text .fragment}

The second derivatives of the log likelihood are
$$
\frac{\partial^2}{\partial\mu_{k\ell}^2} \ell(\boldsymbol{\mu}) 
=-\frac{n_{k\ell}}{\mu_{k\ell}^2}  
$$
and all other second derivatives, namely $\frac{\partial^2}{\partial\mu_{k\ell}\partial\mu_{ij}} \ell(\boldsymbol{\mu})$, are zero.

<div style="margin-top: 30pt;"></div>

:::{.fragment}
Therefore, we have that the MLEs are approximately independent (in large samples) and
$$
\text{Var}\left(\hat\mu_{k\ell}\right)=\frac{\mu_{k\ell}^2}{n_{k\ell}}
\qquad\Rightarrow\qquad
\text{se}\left(\hat\mu_{k\ell}\right)=\frac{\hat\mu_{k\ell}}{n_{k\ell}^{1/2}}
$$
so [**confidence intervals**]{.secondary} for the transition intensities can be easily constructed, based on the large sample [**normal distribution of the MLE**]{.alert}.

:::
:::

## Example [**(Gap on page 111)**]{.alert}

:::{.justify .small-text}

Let's get back to the numerical example we introduced in the beginning of the chapter.


| [Transition]{.alert} | Start | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | End |
| :--- | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: | :---: |
| **Time** | 0.3 | 0.32 | 0.41 | 0.56 | 0.82 | 0.84 | 1.31 | 1.41 | 1.55 | 1.79 | 2.04 | 2.24 | 2.80 |
| **State** | 2 | 1 | 3 | 4 | 1 | 2 | 1 | 3 | 2 | 3 | 1 | 2 | 2 |


```{r, echo = FALSE, fig.width = 10, fig.height = 3.5, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5, 5, 3, 3))

trans<-0:12
state<-c(2,1,3,4,1,2,1,3,2,3,1,2,2)
time<-c(0.3,0.32,0.41,0.56,0.82,0.84,1.31, 1.41,1.55,1.79,2.04,2.24,2.80)

plot(0,0,xlab="Time",ylab="State",axes = FALSE, type="n", xlim=c(0,3), ylim=c(0.5,4.5))
axis(side=2,at=1:4,labels=c("1","2","3","4"),las=2)
axis(side=1,at=(0:6)/2)
for(i in 1:12){
  segments(x0=time[i],x1=time[i+1],y0=state[i],lwd=2, col = "red")
}
```

<div style="margin-top: 20pt;"></div>

:::{.fragment}

$$
t^+_1=0.41,\quad t^+_2=1.29,\quad t^+_3=0.54, ~~~\text{and}~~~~ t^+_4=0.26.
$$

:::

:::


## Example [**(Gap on page 111)**]{.alert}

:::{.justify .small-text}

$$
[n_{k\ell}]=\left(\begin{array}{cccc}
-&2&2&0\\
2&-&1&0\\
1&1&-&1\\
1&0&0&-
\end{array}\right),
$$

$$
[\hat\mu_{k\ell}]=\left(\begin{array}{cccc}
-&\frac{2}{0.41}&\frac{2}{0.41}&0\\[2pt]
\frac{2}{1.29}&-&\frac{1}{1.29}&0\\[2pt]
\frac{1}{0.54}&\frac{1}{0.54}&-&\frac{1}{0.54}\\[2pt]
\frac{1}{0.26}&0&0&-
\end{array}\right)=
\left(\begin{array}{cccc}
-&4.88&4.88&0\\
1.55&-&0.78&0\\
1.85&1.85&-&1.85\\
3.85&0&0&-
\end{array}\right), ~~~\text{and}
$$

$$
[\text{se}(\hat\mu_{k\ell})]=\left(\begin{array}{cccc}
-&3.45&3.45&0\\
1.10&-&0.78&0\\
1.85&1.85&-&1.85\\
3.85&0&0&-
\end{array}\right).
$$

:::

## Multiple sequences $(n > 1)$ [**(Gap on page 112)**]{.alert}

:::{.justify .small-text}

In practice, we will observe [**multiple histories of transitions**]{.secondary} corresponding to individuals $i=1,\cdots ,n$ in the sample, with transition frequencies $[n_{ik\ell}]$ and holding times $\{t^+_{ik}\}$.

:::{.fragment}

We make the standard assumption that the $n$ observed histories are realizations of [**independent**]{.alert} random processes, so  the [**full likelihood (joint probability)**]{.alert} is obtained by multiplying the individual likelihoods
$$
\begin{eqnarray*}
L(\boldsymbol{\mu})&=&\prod_{i=1}^n \;\prod_{k,\ell:k\ne \ell}\mu_{k\ell}^{n_{ik\ell}}\exp\left(-t^+_{ik} \mu_{k\ell}
\right) \\
&=&\prod_{k,\ell:k\ne \ell}\mu_{k\ell}^{n_{k\ell}}\exp\left(-t^+_k \mu_{k\ell}\right) 
\end{eqnarray*}
$$

where now

- $t^+_{k}$ is the total observed holding time in state $k$ [**across all individuals**]{.alert} and 
- $n_{k\ell}$ is the total observed number of transitions from state $k$ to state $\ell$ [**across all individuals**]{.alert}.

:::

<div style="margin-top: -5pt;"></div>

:::{.fragment}

Inference proceeds exactly as [**Sections 9.4**]{.secondary} and [**9.5**]{.secondary} with these new sufficient statistics.

:::

:::



