---
title: "Survival Models (MATH3085/6143)"
subtitle: "Chapter 7: Survival Regression Models"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    progress: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: "André Victor Ribeiro Amaral"
    url: "https://www.avramaral.com/"  
    email: "a.v.ribeiro-amaral@soton.ac.uk"
    affiliations: "University of Southampton"
date: last-modified
date-format: "DD/MM/YYYY"
---

```{r setup}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
knitr::opts_chunk$set(include = TRUE)
options(digits = 4)
library("tidyverse")
library("MASS")
library("survival")
```

## Recap

:::{.justify}

In the last chapter, we

- [**Defined the non-parametric MLE:**]{.alert} we resolved the problem of the unbounded likelihood by constructing a the Non-Parametric Maximum Likelihood Estimator of the survival distribution based on a discrete distribution with mass only at the observed failure times.

- Derived the [**Kaplan-Meier estimator**]{.alert} for the survival function $\hat{S}(t)$.

- Learned how to calculate the variance using Greenwood's formula and how to use the [Nelson-Aalen estimator $\hat{H}(t)$]{.alert} and its resulting [Fleming-Harrington survival curve]{.alert} as alternatives to Kaplan-Meier.

:::

# <span style="font-size: 42px; color: #131516; display: block; margin-bottom: -50px">[**Chapter 7:**]{.secondary} Survival Regression Models</span>


## Introduction

:::{.justify .small-text}

Commonly, when we observe (possibly censored) survival times $t_1, \cdots ,t_n$, we also observe the values of $k$ other variables, $x_1, \cdots ,x_k$, for each of the $n$ units of observation. 

<div style="margin-top: 20pt;"></div>

:::{.fragment}

Then, we drop the assumption that the survival time variables $T_1, \cdots ,T_n$ are identically distributed, and investigate how their distribution  depends on  the [**explanatory variables**]{.alert} (or [**covariates**]{.alert}) $x_1, \cdots, x_k$.

<div style="margin-top: 20pt;"></div>

:::

:::{.fragment}

In a [**regression model**]{.alert}, we assume that the dependence of the distribution of $T_i$ on the values of $x_1, \cdots, x_k$ is through a regression function, which is typically assumed to have linear structure, as
$$
\eta_i=\beta_1 x_{i1}+\beta_2 x_{i2}+ \cdots +\beta_k x_{ik}=\sum_{j=1}^k x_{ij}\beta_j=\mathbf{x}_i^{\top}\boldsymbol{\beta}
$$
where $\mathbf{x}_i=(x_{i1},x_{i2},\ldots ,x_{ik})^{\top}$ is the $k$-vector containing the values of $x_1,\cdots,x_k$ for unit $i$ and $\boldsymbol{\beta}=(\beta_{1},\beta_{2},\ldots ,\beta_{k})^{\top}$ is a $k$-vector of regression parameters.

:::
:::


## Example: Leukaemia (as in [Chapter 2]{.secondary})

:::{.justify}

Survival times are given for 33 patients who died from [acute myelogenous leukaemia]{.alert}. In `R`, this can be found in the `leuk` object in the `MASS` package.

```{r, echo = TRUE, fig.width = 5, fig.height = 3.75, fig.align = "center"}
as_tibble(leuk)
```

`leuk` is a data frame with columns: `wbc` (white blood count), `ag` (a test result, "present" or "absent"), and `time` (survival time in weeks).

:::

## Example: Leukaemia (as in [Chapter 2]{.secondary})

:::{.justify .small-text}

Here, we have two potential explanatory variables, `wbc` and `ag`. As `ag` is a factor (a non-numerical variable) we transform it in the regression function to an indicator (dummy) variable or variables, for example
$$
1(\texttt{ag}~~\text{is present}) = 
\left\{ \begin{array}{ll}
1 & \text{if}~~\texttt{ag}~~\text{is present}\\
0 & \text{if}~~\texttt{ag}~~\text{is absent}
\end{array} 
\right.
$$

<div style="margin-top: 40pt;"></div>

:::{.fragment}

Similarly [**further explanatory variables may be created**]{.alert} from the numeric covariate `wbc`, e.g. $\texttt{wbc}^2$ to investigate possible quadratic dependence.

:::
:::

## Proportional hazards [**(Gap on page 59)**]{.alert}

:::{.justify .small-text}

As was the case with homogeneous survival models in [**Chapter 6**]{.secondary}, we frequently want to investigate the dependence of $T_i$ on $x_1, \cdots, x_k$, [**without**]{.alert} assuming a particular parametric family for $f_{T_i}$.

:::{.fragment}

<div style="margin-top: 60pt;"></div>

A [**proportional hazards**]{.alert} model (or Cox regression model) assumes that the hazard function for the survival variable corresponding to the $i$-th unit is
$$
h_{T_i}(t)=\exp\left\{\mathbf{x}_i^{\top}\boldsymbol{\beta}\right\}\cdot h_0(t)
$$
where $h_0$ is called the [**baseline hazard function**]{.alert} and is not assumed to have a particular mathematical form.

:::

:::{.fragment}

<div style="margin-top: 60pt;"></div>

Hence, we model the effect of the explanatory variables on the distribution of $T_i$ as multiplying the baseline hazard by the exponentiated regression function, $\exp\{\eta_i\}$.

:::
:::

## Proportional hazards [**(Gap on page 60)**]{.alert}

:::{.justify .small-text}

Equivalently, 
$$
\begin{eqnarray*}
S_{T_i}(t) & = & \exp\left[ - \int_0^t h_{T_i}(s) \mathrm{d}s \right]\\
& = & \exp\left[ - \int_0^t \exp\left(\mathbf{x}_i^T\boldsymbol{\beta}\right) \cdot h_0(s)\mathrm{d}s \right]\\
& = & \exp\left[ - \int_0^t h_0(s)\mathrm{d}s \right]^{\exp\left(\mathbf{x}_i^{\top}\boldsymbol{\beta}\right)}\\
& = & S_0(t)^{\exp\left(\mathbf{x}_i^T\boldsymbol{\beta}\right)}
\end{eqnarray*}
$$
where
$$S_0(t) = \exp\left[ - \int_0^t h_0(s)\mathrm{d}s \right]$$
is the baseline survivor function.
:::

## Partial likelihood [**(Gap on page 61)**]{.alert}

:::{.justify .small-text .fragment}

We require to estimate the regression parameters $\boldsymbol{\beta}$ without specifying (or even estimating) the baseline hazard $h_0$ (so we cannot write down a likelihood, as $f_{T_i}$ and $S_{T_i}$ are not fully specified).

The [**partial likelihood function**]{.alert} is the joint probability of the observed data [**conditional on the observed failure times $\{t_i: d_i=1\}$**]{.secondary}.

Conditioning on $\{t_i: d_i=1\}$, the partial likelihood reduces to the probability of the failures occurring in the order in which they were observed.

Let $R_i$ be the [**risk set**]{.alert} at $t_i$, that is
$$
R_i=\{j:t_j\ge t_i\}
$$
the units with failure or censoring times at, or after, $t_i$. Then, 
$$
\mathbb{P}(T_j\in[t_i,t_i+\delta t]\,|\,j\in R_i)=h_{T_j}(t_i)\delta t=\exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)h_0(t_i)\delta t
$$
for some (infinitesimally small) $\delta t$.
:::

## Partial likelihood [**(Gap on page 61)**]{.alert}

:::{.justify .small-text}

Therefore
$$
\mathbb{P}(\mbox{some } j\in R_i \mbox{ fails in } [t_i,t_i+\delta t])=\sum_{j\in R_i} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)h_0(t_i)\delta t+O(\delta t^2).
$$

Hence,
$$
\begin{eqnarray*}
\mathbb{P}(T_i\in [t_i,t_i+\delta t] \,|\, \mbox{some } j\in R_i \mbox{ fails in } [t_i,t_i+\delta t])&& = \\[3pt]
&&\hspace{-12cm}=\;\frac
{\exp\left(\mathbf{x}_i^T\boldsymbol{\beta}\right)h_0(t_i)\delta t}
{\sum_{j\in R_i} \exp\left(\mathbf{x}_j^T\boldsymbol{\beta}\right)h_0(t_i)\delta t+O(\delta t^2)}=\frac
{\exp\left(\mathbf{x}_i^T\boldsymbol{\beta}\right)}
{\sum_{j\in R_i} \exp\left(\mathbf{x}_j^T\boldsymbol{\beta}\right)+O(\delta t)}\\
\end{eqnarray*}
$$
and in the limit $\delta t\to 0$,
$$
\mathbb{P}(i \mbox{ fails at } t_i \,|\, \mbox{some } j\in R_i \mbox{ fails at } t_i)
=\frac
{\exp\left(\mathbf{x}_i^{\top}\boldsymbol{\beta}\right)}
{\sum_{j\in R_i} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)}.
$$
:::

## Partial likelihood [**(Gap on page 61)**]{.alert}

:::{.justify .small-text}

The [**partial likelihood**]{.alert} is therefore given by
$$
\begin{eqnarray*}
L(\boldsymbol{\beta})&=&\mathbb{P}(\mbox{failures occurred in observed order})\\[3pt]
&=&\prod_{i:d_i=1}
\mathbb{P}(i \mbox{ fails at } t_i \,|\, \mbox{some } j\in R_i \mbox{ fails at } t_i)\\[3pt]
&=&\prod_{i:d_i=1}\frac
{\exp\left(\mathbf{x}_i^{\top}\boldsymbol{\beta}\right)}
{\sum_{j\in R_i} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)},
\end{eqnarray*}
$$
which does not depend on the baseline hazard $h_0$.
:::

## Tied failure times [**(Gap on page 63)**]{.alert}

:::{.justify .small-text .fragment}

Suppose two observations (for example $i=1$ and $i=2$) have the same recorded failure time. Our definition of $R_i=\{j:t_j\ge t_i\}$ includes $1\in R_2$ and $2\in R_1$, [**both of which cannot be true**]{.alert}.

:::{.fragment}

Different partial likelihoods arise from assuming

- $1$ failed just before $2$ (so $2\in R_1$ but $1\not\in R_2$).
- $2$ failed just before $1$ (so $2\not\in R_1$ but $1\in R_2$).

<div style="margin-top: -15pt;"></div>

:::

:::{.fragment}

The correct action is to [**average over these two assumptions**]{.alert}, so we replace the contributions of units $i=1$ and $i=2$ to the partial likelihood with
$$
\frac
{\exp\left(\mathbf{x}_1^{\top}\boldsymbol{\beta}\right)}
{\sum_{j\in R_1} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)}
\frac
{\exp\left(\mathbf{x}_2^{\top}\boldsymbol{\beta}\right)}
{\sum_{j\in R_2'} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)}
+
\frac
{\exp\left(\mathbf{x}_2^{\top}\boldsymbol{\beta}\right)}
{\sum_{j\in R_2} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)}
\frac
{\exp\left(\mathbf{x}_1^{\top}\boldsymbol{\beta}\right)}
{\sum_{j\in R_1'} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)}
$$ 
where $R_1'=R_1\setminus\{2\}$ and $R_2'=R_2\setminus\{1\}$.

:::

:::{.fragment}

The exact partial likelihood for tied failures times can be complicated if we have $m>2$ tied failure times ($m!$ terms in the average). [There exist approximations to the likelihood (e.g. Breslow and Efron) in the case of tied failure times.]{.alert} 
:::
:::

## Estimation 

:::{.justify .small-text}

We estimate the regression parameters $\boldsymbol{\beta}$ using the values $\hat{\boldsymbol{\beta}}$ which [maximise the partial likelihood $L(\boldsymbol{\beta})$, or **partial log-likelihood**]{.alert}
$$
\ell(\boldsymbol{\beta})=\sum_{i:d_i=1} \mathbf{x}_i^{\top}\boldsymbol{\beta}-
\sum_{i:d_i=1}\log\left[\sum_{j\in R_i} \exp\left(\mathbf{x}_j^{\top}\boldsymbol{\beta}\right)\right]
$$

:::{.fragment}

Maximisation is performed by solving (numerically) the simultaneous equations
$$
\frac{\partial}{\partial\beta_i} \ell(\boldsymbol{\beta})=0,\qquad i=1,\ldots ,k.
$$

:::

:::{.fragment}

We also obtain
$$
\text{Var}(\hat\beta_i)\approx [I(\boldsymbol{\beta})^{-1}]_{ii}
$$
where $I(\boldsymbol{\beta})$ is the [**observed information matrix**]{.alert} defined by
$$
I(\boldsymbol{\beta})_{ij}=-\frac{\partial^2}{\partial\beta_i\partial\beta_j} \ell(\boldsymbol{\beta}).
$$
:::
:::

## Confidence intervals [**(Gap on page 64)**]{.alert}

:::{.justify .small-text .fragment}

As with maximum likelihood, [maximum partial likelihood estimators have an asymptotic ($n\to\infty$) Normal distribution]{.alert}, so
$$
\frac{\hat\beta_i-\beta_i}{\text{se}(\hat\beta_i)}\sim \text{Normal}(0,1)
$$
can be used as an approximation in moderately large samples, where
$$
\text{se}(\hat\beta_i)^2= [I(\hat{\boldsymbol{\beta}})^{-1}]_{ii}
$$
is an estimate of the [**(asymptotic) standard deviation of $\hat{\beta_i}$**]{.alert}. 

<div style="margin-top: 10pt;"></div>

:::{.fragment}

Hence, a [**$(1-\alpha)\%$ confidence interval**]{.alert} for $\beta_i$ is
$$
[\hat\beta_i- z_{\left(1 - \frac{\alpha}{2}\right)} \cdot \text{se}(\hat\beta_i),~\hat\beta_i+z_{\left(1 - \frac{\alpha}{2}\right)} \cdot \text{se}(\hat\beta_i)]
$$

<div style="margin-top: 30pt;"></div>
:::

:::{.fragment}

In practice, [we rely on computer packages to compute standard errors.]{.alert}
:::

:::

## Hypothesis testing [**(Gap on page 65)**]{.alert}

:::{.justify .small-text .fragment}

In regression models, the hypothesis $H_0:\beta_j=0$ is equivalent to a model where the explanatory variable $x_j$ has no effect on survival (as its values are omitted from the regression function $\mathbf{x}_i^{\top}\boldsymbol{\beta}$).

<div style="margin-top: -10pt;"></div>

:::{.fragment}
We test this hypothesis, by noting that, under $H_0:\beta_j=0$, we have the approximation
$$
\frac{\hat\beta_j - 0}{\text{se}(\hat\beta_j)} = \frac{\hat\beta_j}{\text{se}(\hat\beta_j)}\sim \text{Normal}(0,1).
$$

<div style="margin-top: -10pt;"></div>

:::

:::{.fragment}
So $\hat\beta_j/\text{se}(\hat\beta_j)$ is a test statistic whose values can be calibrated against a standard normal distribution. [For a $\alpha\%$ significance level, we reject $H_0$]{.alert} when
$$
\left|\frac{\hat\beta_j}{\text{se}(\hat\beta_j)}\right|>z_{\left(1 - \frac{\alpha}{2}\right)}.
$$

<div style="margin-top: -10pt;"></div>

:::

:::{.fragment}

We can also calculate a [**p-value**]{.alert} using (this is known as [**Wald test**]{.alert})
$$
\text{p-value} = 2\cdot \left(1 - \mathbb{P}\left(Z \le \left|\frac{\hat\beta_j}{\text{se}(\hat\beta_j)}\right|\right)\right).
$$


:::
:::

## Estimating the baseline $h_0(t)$, $H_0(t)$ and $S_0(t)$

:::{.justify .small-text}

It is not necessary to estimate $h_0$ to answer interesting questions about which covariates affect survival time, and how they do so.

:::{.fragment}

However, as in the homogeneous model, [we can estimate the complete survival distribution nonparametrically]{.alert}, by estimating the hazard function as if the underlying process was discrete.

:::

:::{.fragment}

The [**total failure probability**]{.alert} (across the whole risk set) at failure time $t_i$ is
$$
\sum_{j\in R_i} h_{T_j}(t_i) \quad=\quad\sum_{j\in R_i} \exp(\mathbf{x}_j^{\top}\boldsymbol{\beta}) \cdot h_0(t_i) 
$$
<!--
A Poisson approximation to the distribution of the number of failures at this time produces the estimator
$$
\sum_{j\in R_i} \exp(\mathbf{x}_j^{\top}\hat{\boldsymbol{\beta}}) \cdot \hat h_0(t_i) =d'_i \qquad\Rightarrow\qquad \hat h_0(t_i) =\frac{d'_i}{\sum_{j\in R_i} \exp(\mathbf{x}_j^{\top}\hat{\boldsymbol{\beta}})}
$$
where $d'_i$ is the total number of failures observed at $t_i$.
-->
:::

:::{.fragment}

An [**estimate of the baseline hazard**]{.alert} is given by 
$$
\hat h_0(t_i) =\frac{d'_i}{\sum_{j\in R_i} \exp(\mathbf{x}_j^{\top}\hat{\boldsymbol{\beta}})},
$$
where $d'_i$ is the total number of failures observed at $t_i$.

:::
:::


## Estimating the baseline $h_0(t)$, $H_0(t)$ and $S_0(t)$

:::{.justify .small-text}

As in the homogeneous case in [**Chapter 6**]{.secondary}, the discrete hazard estimates can be transformed into an estimate of the cumulative hazard or survival functions. 

:::{.fragment}

Again, we let $t'_1,\cdots ,t'_m$ be the $m$ ordered distinct failure times with corresponding numbers of failures $d'_1,\cdots ,d'_m$. Then
$$
\hat{H}_0(t)=\sum_{j=1}^i \hat h_0(t'_j)=\sum_{j=1}^i \frac{d'_j}{\sum_{k:t_k\ge t'_j} \exp(\mathbf{x}_k^{\top}\hat{\boldsymbol{\beta}})}\quad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots ,m
$$

:::

:::{.fragment}

(like the Nelson-Aalen estimator in the homogeneous case) and 
$$
\hat{S}_0(t)= \prod_{j=1}^i\exp\left(-\frac{d'_j}{\sum_{k:t_k\ge t'_j} \exp(\mathbf{x}_k^{\top}\hat{\boldsymbol{\beta}})}\right)\qquad t\in[t'_i,t'_{i+1}), \quad i=0,\cdots ,m.
$$
:::

:::{.fragment}

which implies
$$
\hat S_{T_i}(t)=\hat S_0(t)^{\exp\left(\mathbf{x}_i^{\top}{\hat{\boldsymbol{\beta}}}\right)}.
$$
:::
:::

## Fitting Cox models using `R` (`gehan`)

:::{.justify .small-text}

[**Gehan data**]{.alert}: remission times (in weeks) from a clinical trial of 42 leukaemia patients. Patients matched in pairs and randomised to 6-mercaptopurine (drug) or control (from `MASS` package).

```{r, echo = TRUE, fig.width = 6, fig.height = 2.25, fig.align = "center"}
as_tibble(gehan)
```

:::


## Fitting Cox models using `R` (`gehan`)

:::{.justify .small-text}

Using the `coxph()` and `Surv()` from the `survival` package, we can fit a Cox regression model for `gehan` as follows (using `treat` as covariate)

```{r, echo = TRUE, fig.width = 6, fig.height = 2.25, fig.align = "center"}
gehan.cox <- coxph(Surv(time, event = cens) ~ treat, data = gehan)
summary(gehan.cox)
```

:::

## Fitting Cox models using `R` (`gehan`) [**(Gap on page 68)**]{.alert}

:::{.justify .small-text .fragment}

The p-value associated with the treatment dummy variable is [$1.4 \times 10^{-4}$]{.alert}, i.e. the treatment significantly affects survival time. 

The estimated coefficient for treatment is [$1.57213$]{.alert}. This means the control treatment is estimated to increase the hazard by a factor of 
[
$$
\exp \left(1.57213\right) = 4.81687
$$
]{.alert}
relative to the 6-MP treatment. This estimate can be seen in the `R` output in the previous slide (along with a $95\%$ confidence interval). 

:::


## Fitting Cox models using `R` (`gehan`)

:::{.justify .small-text}

The `R` code below create survival curves. The argument `newdata` allows us to specify values of the explanatory variables. In this case there is one explanatory variable which can take two values. So we compute two survival curves for those two values. For comparison, the [Kaplan Meier survival curves are also plotted]{.alert} in grey.

```{r, echo = TRUE, fig.width = 6, fig.height = 3.4, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5.1, 4.1, 0.5, 2.1))

gehan.S  <- survfit(gehan.cox, newdata = data.frame(treat = c("control", "6-MP")))
gehan.km <- survfit(Surv(time, cens) ~ treat, data = gehan)

plot(gehan.S, ylab = "Survival function", xlab = "Time", lty = c(2,1))
lines(gehan.km, lty = c(1,2), col = 8)
legend("topright", lty = c(1, 2, 1, 2), col = c(1, 1, 8, 8), 
legend = c("6-MP (Cox)","Control (Cox)","6-MP (KM)","Control (KM)"), cex = 0.8)
```

:::

## Fitting Cox models using `R` (`leuk`)

:::{.justify .small-text}
[**Leukaemia data**]{.alert}: survival times are given for 33 patients who died from [acute myelogenous leukaemia]{.alert}. In `R`, this can be found in the `leuk` object in the `MASS` package.

```{r, echo = TRUE, fig.width = 6, fig.height = 2.25, fig.align = "center"}
as_tibble(leuk)
```
:::

## Fitting Cox models using `R` (`leuk`)

:::{.justify .small-text}

Using the `coxph()` and `Surv()` from the `survival` package, we can fit a Cox regression model for `leuk` as follows (using a dummy variable for `ag` and a natural logarithm of `wbc` as covariates)

```{r, echo = TRUE, fig.width = 6, fig.height = 2.25, fig.align = "center"}
leuk.cox <- coxph(Surv(time) ~ ag + log(wbc), data = leuk)
summary(leuk.cox)
```

:::

## Fitting Cox models using `R` (`leuk`) [**(Gap on page 70)**]{.alert}

:::{.justify .small-text .fragment}

The significance of both explanatory variables is immediately obvious from the two p-values [$0.01276$]{.alert} and [$0.0068$]{.alert}. Strictly, we should formal model selection using either backwards or forwards selection, or using information criteria (see `MATH2010`).

The estimated coefficient for [`ag` is $-1.06905$]{.alert}. This means `ag` being present is estimated to increase the hazard by a factor of 
[
$$
\exp \left(-1.06905\right) = 0.34333,
$$
]{.alert}
relative to `ag` being absent, i.e. `ag` being present reduces the hazard. 

The estimated coefficient for [`log(wbc)` is $0.3677$]{.alert}. This $\log(\text{wbc})$ increasing by one unit is estimated to increase the hazard by a factor of 
[
$$
\exp\left(0.3677\right) = 1.44441,
$$
]{.alert}
i.e. the higher the `wbc`, the higher the hazard. 
:::


## Fitting Cox models using `R` (`leuk`)

:::{.justify .small-text}

Similarly to before, we can plot the estimated survival functions as follows.

```{r, echo = TRUE, fig.width = 6, fig.height = 3.4, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5.1, 4.1, 0.5, 2.1))

leuk.S <- survfit(leuk.cox, newdata = data.frame(ag  = c("present", "present", "absent", "absent"), 
                                                 wbc = c(750, 100000, 750, 100000)))

plot(leuk.S, ylab= "Survival function", xlab = "Time", lty = c(1, 2, 1, 2), col = c(1, 1, 8, 8))
legend("topright", lty = c(1, 2, 1, 2), col = c(1, 1, 8, 8), 
legend = c("Present (750)", "Present (100,000)","Absent  (750)", "Absent  (100,000)"), cex = 0.8)
```

:::

