---
title: "Survival Models (MATH3085/6143)"
subtitle: "Chapter 8: Multistate Survival Models"
format:
  revealjs: 
    theme: [default, custom.scss]
    slide-number: true
    progress: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: "André Victor Ribeiro Amaral"
    url: "https://www.avramaral.com/"  
    email: "a.v.ribeiro-amaral@soton.ac.uk"
    affiliations: "University of Southampton"
date: last-modified
date-format: "DD/MM/YYYY"
---

```{r setup}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
knitr::opts_chunk$set(include = TRUE)
options(digits = 4)
library("tidyverse")
```

## Recap

:::{.justify}

In the last chapter, we

- Defined the semi-parametric [**Cox PH model**]{.alert}, where covariates multiply the baseline hazard $h_0(t)$

- Established that the regression parameters $\boldsymbol{\beta}$ are estimated by maximising the [**partial likelihood**]{.alert}, which conditions out the unknown baseline hazard.

- Introduced [**Accelerated Failure Time**]{.alert} (AFT) models as the fully parametric alternative where covariates shrink or stretch the time axis.

:::

# <span style="font-size: 42px; color: #131516; display: block; margin-bottom: -50px">[**Chapter 8:**]{.secondary} Multistate Survival Models</span>

## Introduction

:::{.justify .small-text}

In [**Chapters 1 to 7**]{.secondary}, we have assumed that survival (time-to-failure) of each unit is an observation of a random variable $T$, and our survival models have been stochastic (probability) models describing the distribution of $T$.


:::{.fragment}

An alternative approach is to model the variable $Y_t$ representing the [**status (alive or dead)**]{.alert} of a unit at time $t$.

:::

:::

## Introduction

:::{.justify .small-text}

A probability model for a time-indexed variable (one which changes over time) is  a [**stochastic process**]{.alert}. 

`survival` is a simple stochastic process where, at time $t=0$, $Y_0=1$ (`alive`) and then the process remains in state $1$ unless at some value of $t$ a transition to $Y_t=2$ (`dead`) is made after which the process remains in state $2$.

<div style="margin-top: 40pt;"></div>

```{r, echo = FALSE, fig.width = 10, fig.height = 3.75, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(5.1, 4.1, 0.5, 2.1))

diff<-1.78
plot(0,0,type="n",xlab="Time",ylab="State",xlim=c(0,3),ylim=c(1,2),axes=FALSE)
segments(x0=0,x1=diff,y0=1,y1=1,lwd=2)
segments(x0=diff,x1=diff, y0=1,y1=2,lwd=2)
arrows(x0=diff,x1=3,y0=2,y1=2,len=0.1,lwd=2)
axis(side=2,at=c(1,2),labels=c("1","2"),las=2)
axis(side=1,at=(0:6)/2)
```

:::

## State space [**(Gap on page 82)**]{.alert}


:::{.justify .small-text}

The set of possible values that a stochastic process $Y_t$ can take over time $t$ is called its [**state space**]{.alert}, $S$.


:::{.fragment}

In the simple survival model $S=\{\mbox{alive},\mbox{dead}\}$ or $S=\{1,2\}$.

This is an example of a [discrete state space]{.alert}, and in this module we will restrict consideration to discrete state spaces with a small number of possible states.

The simple survival state space can be represented as

<div style="margin-top: -20pt;"></div>


```{r, echo = FALSE, fig.width = 10, fig.height = 2, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(0, 0, 0, 0))

plot(0,0,xlim=c(0,100),ylim=c(0,40),type="n",axes=FALSE,xlab="",ylab="")

polygon(x=c(20,20,40,40),y=c(10,30,30,10))
polygon(x=c(20,20,40,40)+40,y=c(10,30,30,10))
arrows(x0=40,x1=60,y0=20,y1=20,len=0.1)
text(x=30,y=20,labels=c("Alive"),pos=3)
text(x=30,y=20,labels=c("(Y=1)"),pos=1)
text(x=30+40,y=20,labels=c("Dead"),pos=3)
text(x=30+40,y=20,labels=c("(Y=2)"),pos=1)
```

<div style="margin-top: -50pt;"></div>

The arrow indicates that transition is possible only in one direction.

In this case the state $\mbox{dead}$ is called an [**absorbing**]{.secondary} state.

:::

:::

## Multi-state models [**(Gap on page 83)**]{.alert}


:::{.justify .small-text}

Stochastic processes allow us to model richer survival time processes than simply two-state $\{\text{alive}, ~\text{dead}\}$ processes with an absorbing ($\text{dead}$) state.

:::{.fragment}

Two such examples are:

**General two-state model (no absorbing states)**

<div style="margin-top: -15pt;"></div>

```{r, echo = FALSE, fig.width = 10, fig.height = 2, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(0, 0, 0, 0))

plot(0,0,xlim=c(0,100),ylim=c(0,40),type="n",axes=FALSE,xlab="",ylab="")
polygon(x=c(20,20,40,40),y=c(10,30,30,10))
polygon(x=c(20,20,40,40)+40,y=c(10,30,30,10))
arrows(x0=40,x1=60,y0=15,y1=15,len=0.1)
arrows(x0=60,x1=40,y0=25,y1=25,len=0.1)
text(x=30,y=20,labels=c("Alive"),pos=3)
text(x=30,y=20,labels=c("(Y=1)"),pos=1)
text(x=30+40,y=20,labels=c("Ill"),pos=3)
text(x=30+40,y=20,labels=c("(Y=2)"),pos=1)
```

<div style="margin-top: -50pt;"></div>

:::

:::{.fragment}

**Multiple decrement (absorbing state) model**

```{r, echo = FALSE, fig.width = 10, fig.height = 2, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(0, 0, 0, 0))

plot(0,0,xlim=c(0,100),ylim=c(0,70),type="n",axes=FALSE,xlab="",ylab="")
polygon(x=c(20,20,40,40),y=c(25,45,45,25))
polygon(x=c(60,60,80,80),y=c(0,20,20,0))
polygon(x=c(60,60,80,80),y=c(0,20,20,0)+25)
polygon(x=c(60,60,80,80),y=c(0,20,20,0)+50)
arrows(x0=40,x1=60,y0=30,y1=10,len=0.1)
arrows(x0=40,x1=60,y0=35,y1=35,len=0.1)
arrows(x0=40,x1=60,y0=45,y1=60,len=0.1)
text(x=30,y=35,labels=c("Alive"),pos=3)
text(x=30,y=35,labels=c("(Y=1)"),pos=1)
text(x=70,y=63.66,labels=c("Dead"),pos=3,cex=0.8,offset=0.25)
text(x=70,y=63.66,labels=c("Cancer"),pos=1,cex=0.8,offset=0.25)
text(x=70,y=56.66,labels=c("(Y=3)"),pos=1,cex=0.8,offset=0.25)
text(x=70,y=63.66-25,labels=c("Dead"),pos=3,cex=0.8,offset=0.25)
text(x=70,y=63.66-25,labels=c("Coronary Heart Disease"),pos=1,cex=0.8,offset=0.25)
text(x=70,y=56.66-25,labels=c("(Y=2)"),pos=1,cex=0.8,offset=0.25)
text(x=70,y=63.66-50,labels=c("Dead"),pos=3,cex=0.8,offset=0.25)
text(x=70,y=63.66-50,labels=c("Other"),pos=1,cex=0.8,offset=0.25)
text(x=70,y=56.66-50,labels=c("(Y=4)"),pos=1,cex=0.8,offset=0.25)
```

:::

:::

## Markov processes

:::{.justify .small-text}

A [**Markov process**]{.alert} model for $Y_t$ is one where the future is [conditionally independent of the history of the process given the current state.]{.secondary} For example, tomorrow's weather only depends on today's weather, and is independent on all weather before today. 

<div style="margin-top: 35pt;"></div>

:::{.fragment}

Hence, a Markov process can be specified by [**transition probabilities**]{.alert}
$$
\mathbb{P}(Y_{x+t}=j\,|\,Y_x=i)\;\equiv\;p_{ij}(x,t),\quad j\in S
$$
for any present time $x$, future time $x+t$ and present state $i$. This makes explicit that the probability, at time $x$, of future realisations of the process depends only on the current state $Y_x$, and not on the [**history**]{.secondary} of $Y$ (its values in the period $[0,t)$).

:::

<div style="margin-top: 35pt;"></div>

:::{.fragment}

For a [**time-homogeneous**]{.alert} Markov process, the transition probabilities do not depend on the value of $x$, so we can write
$$
P(Y_{x+t}=j\,|\,Y_x=i)\;=\;p_{ij}(t).
$$

:::

:::


## Transition intensity [**(Gap on page 84)**]{.alert}

:::{.justify .small-text .fragment}

The [**transition intensity function $\mu_{ij}(x)$**]{.alert} is defined as

$$
\mu_{ij}(x)=\lim_{\delta t\to 0} \frac{p_{ij}(x,\delta t)}{\delta t}, \qquad i\ne j.
$$
<div style="margin-top: -5pt;"></div>

It follows that

$$
p_{ij}(x,\delta t)=\mu_{ij}(x)\delta t +o(\delta t), \qquad i\ne j.
$$
<div style="margin-top: -5pt;"></div>

We define

$$
\mu_{ii}(x)=-\sum_{j\ne i} \mu_{ij}(x),
$$
<div style="margin-top: -5pt;"></div>

so that 

$$
p_{ii}(x,\delta t)=1+\mu_{ii}(x) \cdot \delta t +o(\delta t), \qquad i\ne j.
$$
<!--
In Markov process theory, we often use $q$ to denote transition intensity, but $\mu$ is more common in survival applications.
-->

:::{.fragment}

For any transition $i \to j$ which is prohibited (e.g. $\mbox{dead}\to\mbox{alive}$), we have $\mu_{ij}=0$.

For a [**homogenous process**]{.secondary}, transition intensity $\mu_{ij}(x)$ does not depend on $x$, i.e. it is written $\mu_{ij}$.

:::

::::


## Chapman-Kolmogorov equations: solving a Markov process for $p_{ij}(x,t)$

:::{.justify .small-text}

The transition intensity function $\mu_{ij}(x)$ (or just $\mu_{ij}$ for a time-homogeneous process) provides an efficient representation of the process.

<div style="margin-top: 30pt;"></div>

:::{.fragment}
[How are the transition probabilities $p_{ij}(x,t)$ obtained from the transition intensities $\mu_{ij}(x)$?]{.secondary}

:::

<div style="margin-top: 30pt;"></div>

:::{.fragment}

We need to derive the [**Kolmogorov forward equations**]{.alert}, the solution to which are the required transition probabilities $p_{ij}(x,t)$.

:::

:::

## Chapman-Kolmogorov equations: solving a Markov process for $p_{ij}(x,t)$ [**(Gap on page 85)**]{.alert}

:::{.justify .small-text}

For any times $x,s,t\ge 0$ and states $i,j,k\in S$, we have

<div style="margin-top: -10pt;"></div>

$$
\begin{eqnarray*}
\mathbb{P}(Y_{x+s+t}=j,Y_{x+s}=k\,|\,Y_x=i)&&\\
&&\hspace{-50mm}=\mathbb{P}(Y_{x+s+t}=j\,|\,Y_{x+s}=k,Y_x=i) \cdot \mathbb{P}(Y_{x+s}=k\,|\,Y_x=i)\\
&&\hspace{-50mm}=\mathbb{P}(Y_{x+s+t}=j\,|\,Y_{x+s}=k) \cdot \mathbb{P}(Y_{x+s}=k\,|\,Y_x=i), \\
\end{eqnarray*}
$$

<div style="margin-top: -10pt;"></div>

Hence

<div style="margin-top: -10pt;"></div>

$$
\begin{eqnarray*}
\mathbb{P}(Y_{x+s+t}=j\,|\,Y_x=i)&&\\
&&\hspace{-20mm}=\sum_{k\in S} \mathbb{P}(Y_{x+s+t}=j,Y_{x+s}=k\,|\,Y_x=i)\\
&&\hspace{-20mm}=\sum_{k\in S} \mathbb{P}(Y_{x+s+t}=j\,|\,Y_{x+s}=k) \cdot \mathbb{P}(Y_{x+s}=k\,|\,Y_x=i), \\
\end{eqnarray*}
$$

<div style="margin-top: -10pt;"></div>

or

<div style="margin-top: -10pt;"></div>


$$
p_{ij}(x,s+t)=\sum_{k\in S} p_{ik}(x,s)p_{kj}(x+s,t), \qquad i,j\in S.
$$

<div style="margin-top: -10pt;"></div>

These are the [**Chapman-Kolmogorov equation**]{.alert}.


:::

## Chapman-Kolmogorov equations: solving a Markov process for $p_{ij}(x,t)$ [**(Gap on page 86)**]{.alert}

:::{.justify .small-text}

Suppose that there are $m$ states in $S$, labelled $1,\cdots ,m$.

For a Markov process, we define the [transition matrix between times $x$ and $x+t$]{.alert} to be the matrix
$$
\text{P}(x,t)=\left(
\begin{array} {cccc}
p_{11}(x,t)&p_{12}(x,t)&\cdots&p_{1m}(x,t)\\
p_{21}(x,t)&p_{22}(x,t)&\cdots&p_{2m}(x,t)\\
\vdots&\vdots&&\vdots\\
p_{m1}(x,t)&p_{m2}(x,t)&\cdots&p_{mm}(x,t)
\end{array}
\right)
$$

<div style="margin-top: 20pt;"></div>

The Chapman-Kolmogorov equations can be written as
$$
\mathbb{P}(x,s+t)=\mathbb{P}(x,s)\cdot \mathbb{P}(x+s,t)
$$

<div style="margin-top: 20pt;"></div>

For a time-homogeneous process, we can drop the first argument $x$ in the above.

:::

## Chapman-Kolmogorov equations: solving a Markov process for $p_{ij}(x,t)$ [**(Gap on page 87)**]{.alert}

:::{.justify .small-text}

The Chapman-Kolmogorov equations give
$$
p_{ij}(x,t+\delta t)=\sum_{k\in S} p_{ik}(x,t)p_{kj}(x+t,\delta t)\qquad i,j\in S
$$
which implies that 
$$
\begin{eqnarray*}
\frac{p_{ij}(x,t+\delta t)-p_{ij}(x,t)}{\delta t}&&\\
&&\hspace{-20mm}=\sum_{k\in S} p_{ik}(x,t) \cdot \frac{p_{kj}(x+t,\delta t)}{\delta t}-\frac{p_{ij}(x,t)}{\delta t}
\\
&&\hspace{-20mm}=\sum_{k\ne j} p_{ik}(x,t) \cdot \frac{p_{kj}(x+t,\delta t)}{\delta t}-p_{ij}(x,t) \cdot \frac{1-p_{jj}(x+t,
\delta t)}{\delta t}.
\\
\end{eqnarray*}
$$



:::


## Chapman-Kolmogorov equations: solving a Markov process for $p_{ij}(x,t)$ [**(Gap on page 88)**]{.alert}

:::{.justify .small-text}

In the limit $\delta t\to 0$, we have
$$
\begin{eqnarray*}
\frac{d}{dt} p_{ij}(x,t)&=&
\sum_{k\ne j} p_{ik}(x,t)\mu_{kj}(x+t)+p_{ij}(x,t) \cdot \mu_{jj}(x+t) \\
&=& \sum_{k\in S} p_{ik}(x,t) \cdot \mu_{kj}(x+t), \qquad i,j\in S\\
\end{eqnarray*}
$$
These are the [**Kolmogorov forward equations**]{.alert}, which are a set of differential equations for $p_{ij}(x,t)$

They can be written in matrix form as $\frac{d}{dt} \text{P}(x,t)= \text{P}(x,t) \cdot \text{M}(x+t)$,
where $\text{M}(x)$ is the transition intensity matrix
$$
\text{M}(x)=\left(
\begin{array} {cccc}
\mu_{11}(x)&\mu_{12}(x)&\cdots&\mu_{1m}(x)\\
\mu_{21}(x)&\mu_{22}(x)&\cdots&\mu_{2m}(x)\\
\vdots&\vdots&&\vdots\\
\mu_{m1}(x)&\mu_{m2}(x)&\cdots&\mu_{mm}(x)
\end{array}
\right).
$$



:::

## The holding time distribution [**(Gap on page 89)**]{.alert}

:::{.justify .small-text}

Consider a unit in state $i$ at time $x$. The holding time $T_{xi}$ is the random variable indicating the length of the time interval between $x$ and the next transition to any other state. 

:::{.fragment}

We have
$$
\begin{eqnarray*}
\mathbb{P}(T_{xi}\le t+\delta t\,|\,T_{xi}\ge t)&=& \mathbb{P}(Y_{x+t+\delta t}\ne i \,|\,Y_{[x,x+t]}=i)\\
&=&1-\mathbb{P}(Y_{x+t+\delta t}= i \,|\,Y_{x+t}=i)
\end{eqnarray*}
$$
and, therefore,
$$
\frac{\mathbb{P}(T_{xi}\le t+\delta t\,|\,T_{xi}\ge t)}{\delta t}=
\frac{1-\mathbb{P}(Y_{x+t+\delta t}= i \,|\,Y_{x+t}=i)}{\delta t}
$$
Taking the limit as $\delta t\to 0$, we have
$$
h_{xi}(t)=-\mu_{ii}(x+t)=\sum_{i\ne j} \mu_{ij}(x+t),
$$
where $h_{xi}$ is the hazard function for the holding time variable $T_{xi}$.

:::
:::

## The holding time distribution [**(Gap on page 89)**]{.alert}

:::{.justify .small-text}

The hazard function $h_{xi}(t)$ can be transformed to a survival function in the usual way, so we have
$$
\begin{eqnarray*}
S_{xi}(t)=\exp\left( -\int_0^t h_{xi}(t) \mathrm{d}t \right) &=& \exp\left( \int_0^t \mu_{ii}(x+t)\mathrm{d}t \right) \\
&=& \exp\left( -\int_0^t \sum_{i\ne j} \mu_{ij}(x+t)\mathrm{d}t \right)
\end{eqnarray*}
$$

<div style="margin-top: 40pt;"></div>

For a time-homogeneous Markov process,  all $\mu_{ij}(x)$ are constant (independent of $x$) so we have $h_{xi}(t)=-\mu_{ii}=\sum_{j\ne i} \mu_{ij}$, a constant hazard. Hence,
$$
S_{xi}(t)= \exp\left(  \mu_{ii} t \right)= \exp\left(  -t \sum_{j\ne i} \mu_{ij}\right)
$$
and the holding time distribution is exponential, with rate
$\sum_{i\ne j} \mu_{ij}$.

:::

## Conditional transition probability [**(Gap on page 90)**]{.alert}

:::{.justify .small-text}

[Conditional on a transition being made from state $i$ at time $x+t$, what is the probability that transition is made to state $j$ for each $j\ne i$?]{.secondary}

:::{.fragment}
$$
\begin{eqnarray*}
&&\mathbb{P}(\mbox{transition from $i$ to $j$ at $x+t$}\;|\;\mbox{transition from $i$ at $x+t$})\\[3pt]
&&\qquad= \lim_{\delta t\to 0}  \mathbb{P}(Y_{x+t+\delta t}=j\;|\;Y_{x+t+\delta t}\ne i \mbox{ and } Y_{x+t}=i)\\[3pt]
&&\qquad= \lim_{\delta t\to 0}  \frac{\mathbb{P}(Y_{x+t+\delta t}=j\;|\;Y_{x+t}= i )}
{\mathbb{P}(Y_{x+t+\delta t}\ne i\;|\; Y_{x+t}=i)}\\[3pt]
&&\qquad= \lim_{\delta t\to 0}  \frac{\mu_{ij}(x+t)\delta t+o(\delta t)}
{\sum_{k\ne i}\mu_{ik}(x+t)\delta t+o(\delta t)}\\[3pt]
&&\qquad=   \frac{\mu_{ij}(x+t)}
{\sum_{k\ne i}\mu_{ik}(x+t)}
\end{eqnarray*}
$$

:::{.fragment}

For a time-homogeneous Markov process, all $\mu_{ij}(x)$ are constant (independent of $x$) so this becomes
$\frac{\mu_{ij}}{\sum_{k\ne i}\mu_{ik}}$.
:::

:::
:::

## Two-state model (absorbing state) [**(Gap on page 91)**]{.alert}

:::{.justify .small-text}

```{r, echo = FALSE, fig.width = 10, fig.height = 2, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(0, 0, 0, 0))

plot(0,0,xlim=c(0,100),ylim=c(0,40),type="n",axes=FALSE,xlab="",ylab="")
polygon(x=c(20,20,40,40),y=c(10,30,30,10))
polygon(x=c(20,20,40,40)+40,y=c(10,30,30,10))
arrows(x0=40,x1=60,y0=20,y1=20,len=0.1)
text(x=30,y=20,labels=c("Alive"),pos=3)
text(x=30,y=20,labels=c("(Y=1)"),pos=1)
text(x=30+40,y=20,labels=c("Dead"),pos=3)
text(x=30+40,y=20,labels=c("(Y=2)"),pos=1)
```

<div style="margin-top: -30pt;"></div>

:::{.fragment}
Any two-state process is determined by the transition intensities between the two states, $\mu_{12}(x)$ and $\mu_{21}(x)$.

As `state 2` is absorbing,
$$
p_{21}(x,t)=0, \quad \mbox{for all }x,t\qquad\Rightarrow\qquad \mu_{21}(x)=0, \quad \mbox{for all }x.
$$

Hence the transition intensity matrix is given by
$$
\text{M}(x)=\left(
\begin{array}{cc}
-\mu_{12}(x)&\mu_{12}(x)\\
0&0
\end{array}
\right).
$$

:::

:::

## Two-state model (absorbing state) [**(Gap on page 91)**]{.alert}

:::{.justify .small-text}

We have 
$$
p_{21}(x,t)=0, \quad \mbox{for all }x,t\qquad\Rightarrow\qquad p_{22}(x,t)=1, \quad \mbox{for all }x,t.
$$

To completely determine the transition structure for the process, all that remains is to solve the Kolmogorov forward equation for either $p_{11}(x,t)$ or $p_{12}(x,t)$, and then obtain the other using
$$
p_{11}(x,t)+p_{12}(x,t)=1
$$
For completeness, we shall solve both and show that they give equivalent results.


:::

## Two-state model (absorbing state) [**(Gap on page 91)**]{.alert}

:::{.justify .small-text}

The Kolmogorov forward equation for $p_{11}(x,t)$ is
$$
\begin{eqnarray*}
\frac{d}{dt} p_{11}(x,t) &=& p_{11}(x,t)\mu_{11}(x+t)+p_{12}(x,t)\mu_{21}(x+t)\\
&=& -p_{11}(x,t)\mu_{12}(x+t).
\end{eqnarray*}
$$
So,
$$
\begin{eqnarray*}
&&\int\frac{\mathrm{d} p_{11}(x,t)}{p_{11}(x,t)} = -\int_0^t\mu_{12}(x+t) \mathrm{d}t\\
\Rightarrow&& \log p_{11}(x,t)= -\int_0^t\mu_{12}(x+t) \mathrm{d}t + C\\
\Rightarrow&& p_{11}(x,t)= A \cdot \exp\left(-\int_0^t\mu_{12}(x+t) \mathrm{d}t \right)\\
\Rightarrow&& p_{11}(x,t)= \exp\left(-\int_0^t\mu_{12}(x+t) \mathrm{d}t \right),
\end{eqnarray*}
$$
because we have the boundary condition $p_{11}(x,0)=1$.

:::

## Two-state model (absorbing state) [**(Gap on page 91)**]{.alert}

:::{.justify .small-text}

The Kolmogorov forward equation for $p_{12}(x,t)$ is
$$
\begin{eqnarray*}
\frac{d}{dt} p_{12}(x,t) &=& p_{11}(x,t)\mu_{12}(x+t)+p_{12}(x,t)\mu_{22}(x+t)\\
&=& p_{11}(x,t)\mu_{12}(x+t) =  [1-p_{12}(x,t)]\cdot \mu_{12}(x+t).\\
\end{eqnarray*}
$$


So,
$$
\begin{eqnarray*}
&&\int\frac{\mathrm{d} p_{12}(x,t)}{1-p_{12}(x,t)} = \int_0^t\mu_{12}(x+t) \mathrm{d}t\\
\Rightarrow&& -\log [1-p_{12}(x,t)]= \int_0^t\mu_{12}(x+t) \mathrm{d}t + C\\
\Rightarrow&& 1-p_{12}(x,t)= A \cdot \exp\left(-\int_0^t\mu_{12}(x+t) \mathrm{d}t \right)\\
\Rightarrow&& p_{12}(x,t)= 1-\exp\left(-\int_0^t\mu_{12}(x+t) \mathrm{d}t \right)
\end{eqnarray*}
$$
because we have the boundary condition $p_{12}(x,0)=0$.
:::

## Two-state model (absorbing state) [**(Gap on page 91)**]{.alert}

:::{.justify .small-text}

Clearly, the two solutions are equivalent, as we have
$$
p_{11}(x,t)= \exp\left(-\int_0^t\mu_{12}(x+t) \mathrm{d}t \right)=1-p_{12}(x,t).
$$
We also note that if we define $T_x$ to be the random variable representing the time (from $x$) in the alive (`state 1`) before death (transition to `state 2`), then
$$
S_{T_x}(t)=p_{11}(x,t)
$$ 
because for this state space, being in `state 1` at time $x+t$ is equivalent to holding in `state 1` throughout the interval $[x,x+t]$.
\medskip

It follows from the solution for $p_{11}(x,t)$ above that
$$
h_{T_x}(t)=- \frac{d}{dt} \log S_{T_x}(t) = -\frac{d}{dt} \log p_{11}(x,t) = \mu_{12}(x+t)
$$ 
so the transition intensity plays the role of the hazard function for the variable representing the time spent in the alive (`state 1`).

:::

## Two-state model (no absorb. state) [**(Gap on page 95)**]{.alert}

:::{.justify .small-text}

```{r, echo = FALSE, fig.width = 10, fig.height = 2, fig.align = "center"}
par(family = "Latin Modern Roman 10", mar = c(0, 0, 0, 0))

plot(0,0,xlim=c(0,100),ylim=c(0,40),type="n",axes=FALSE,xlab="",ylab="")
polygon(x=c(20,20,40,40),y=c(10,30,30,10))
polygon(x=c(20,20,40,40)+40,y=c(10,30,30,10))
arrows(x0=40,x1=60,y0=15,y1=15,len=0.1)
arrows(x0=60,x1=40,y0=25,y1=25,len=0.1)
text(x=30,y=20,labels=c("Alive"),pos=3)
text(x=30,y=20,labels=c("(Y=1)"),pos=1)
text(x=30+40,y=20,labels=c("Ill"),pos=3)
text(x=30+40,y=20,labels=c("(Y=2)"),pos=1)
```

:::{.fragment}

TBD.

:::

:::